---
title: "Bootstrap Inference"
format: html
execute:
  collapse: true
  comment: "#>"
  eval: false
---

## Overview

Bootstrap inference provides confidence intervals for indirect effects without assuming normality. medfit implements three bootstrap methods:

1. **Parametric bootstrap**: Sample from parameter distribution (fast, assumes normality)
2. **Nonparametric bootstrap**: Resample data and refit models (robust, slower)
3. **Plugin estimator**: Point estimate only (fastest, no CI)

All methods return a `BootstrapResult` object with consistent structure.

## Theoretical Background

### Why Bootstrap for Indirect Effects?

The indirect effect in mediation analysis is computed as the product of path coefficients: $ab$. Even when the individual coefficients $a$ and $b$ are normally distributed, their product typically is not (Aroian, 1947; Craig, 1936). This creates challenges for traditional confidence interval methods.

MacKinnon et al. (2002) compared 14 methods for testing mediation and found that:

- Normal-theory methods (Sobel test) have low statistical power
- Bootstrap methods provide the best balance of Type I error and power
- Bias-corrected bootstrap has excellent coverage properties

Preacher & Hayes (2008) further established bootstrap methods as the standard for mediation inference, noting that they:

- Make no assumptions about the shape of the sampling distribution
- Can be applied to complex mediation models
- Provide confidence intervals with accurate coverage rates

## Why Bootstrap for Mediation?

The sampling distribution of indirect effects ($a \times b$) is typically:

- **Non-normal**: Even when a and b are normal, their product is not
- **Skewed**: Often right-skewed
- **Complex**: No closed-form distribution for general case

Bootstrap methods:
- Don't assume normality
- Provide accurate coverage (closer to nominal 95%)
- Handle complex indirect effects (serial mediation, moderated mediation)

## Parametric Bootstrap

Samples from the estimated parameter distribution:

```{r}
#| label: parametric
# Assume we have extracted mediation structure
med <- extract_mediation(model_m, model_y, treatment = "X", mediator = "M")

# Define statistic function for indirect effect
indirect_effect <- function(theta) {
  theta["a"] * theta["b"]
}

# Parametric bootstrap
boot_result <- bootstrap_mediation(
  statistic_fn = indirect_effect,
  method = "parametric",
  mediation_data = med,
  n_boot = 1000,
  ci_level = 0.95,
  seed = 123
)

print(boot_result)
summary(boot_result)
```

### How It Works

The parametric bootstrap assumes that the parameter estimates follow a multivariate normal distribution (Efron & Tibshirani, 1993):

$$\hat{\theta} \sim N(\theta, \Sigma)$$

where $\theta = (a, b, c', \ldots)$ are the mediation path coefficients and $\Sigma$ is their covariance matrix.

**Algorithm:**

1. Extract parameter estimates: $\hat{\theta} = (\hat{a}, \hat{b}, \ldots)$
2. Extract covariance matrix: $\hat{\Sigma}$
3. For each bootstrap iteration $i = 1, \ldots, B$:
   - Sample $\theta^*_i \sim N(\hat{\theta}, \hat{\Sigma})$ using `MASS::mvrnorm()`
   - Compute indirect effect: $IE^*_i = a^*_i \times b^*_i$
4. Compute percentile CI from bootstrap distribution

**Implementation Detail**: When the covariance matrix is near-singular (which can occur with highly correlated predictors), medfit applies eigenvalue adjustment to ensure positive definiteness before sampling.

### When to Use

- **Fast**: No model refitting required
- **Appropriate when**: Parameters are approximately normal (large n)
- **Inappropriate when**: Small samples, non-normal parameters

### Advantages

- Very fast (no refitting)
- Reproducible with seed
- Works with extracted models (no need for original data)

### Disadvantages

- Assumes multivariate normality of parameters
- May underestimate uncertainty in small samples

## Nonparametric Bootstrap

Resamples data and refits models:

```{r}
#| label: nonparametric
# Nonparametric bootstrap (requires original data)
boot_result <- bootstrap_mediation(
  statistic_fn = indirect_effect,
  method = "nonparametric",
  mediation_data = med,
  n_boot = 1000,
  ci_level = 0.95,
  seed = 123
)

print(boot_result)
```

### How It Works

The nonparametric bootstrap (Efron, 1979) makes no distributional assumptions about the data:

**Algorithm:**

1. For each bootstrap iteration $i = 1, \ldots, B$:
   - Resample data with replacement: $D^*_i$ (sample $n$ observations from original $n$)
   - Refit mediator model on $D^*_i$ using the original formula
   - Refit outcome model on $D^*_i$ using the original formula
   - Extract paths: $a^*_i$, $b^*_i$
   - Compute indirect effect: $IE^*_i = a^*_i \times b^*_i$
2. Compute percentile CI from bootstrap distribution

**Implementation Details:**

- Uses the formulas stored in `MediationData` to refit models
- Handles model convergence failures gracefully (returns NA for failed iterations)
- Supports GLM families specified in the original model fit

### When to Use

- **Robust**: Makes no parametric assumptions
- **Appropriate when**: Small samples, non-normal data, GLMs
- **Gold standard**: Most widely accepted method

### Advantages

- No distributional assumptions
- Robust to outliers
- Captures full sampling variability

### Disadvantages

- Computationally intensive (refits models B times)
- Requires original data
- Can be slow for complex models

### Parallel Processing

Speed up with parallel processing:

```{r}
#| label: parallel
# Detect available cores
n_cores <- parallel::detectCores() - 1

# Run in parallel
boot_result <- bootstrap_mediation(
  med,
  method = "nonparametric",
  n_boot = 5000,
  parallel = TRUE,
  n_cores = n_cores
)
```

## Plugin Estimator

Point estimate only, no confidence interval:

```{r}
#| label: plugin
# Plugin estimator (fastest)
plugin_result <- bootstrap_mediation(
  med,
  method = "plugin",
  n_boot = 0 # Not used for plugin
)

print(plugin_result)
```

### When to Use

- Quick checks
- Point estimates for simulation studies
- When CI is not needed

### How It Works

Simply computes $\hat{a} \times \hat{b}$ from the fitted model. No resampling.

## Interpreting Results

### Bootstrap Distribution

The bootstrap distribution shows the sampling variability:

```{r}
#| label: distribution
# Access bootstrap estimates
boot_estimates <- boot_result@boot_estimates

# Histogram
hist(
  boot_estimates,
  breaks = 50,
  main = "Bootstrap Distribution of Indirect Effect",
  xlab = "Indirect Effect (a * b)"
)

# Add percentile CI
abline(v = boot_result@ci_lower, col = "red", lwd = 2)
abline(v = boot_result@ci_upper, col = "red", lwd = 2)
abline(v = boot_result@estimate, col = "blue", lwd = 2)

# Check for normality
qqnorm(boot_estimates)
qqline(boot_estimates)
```

### Confidence Interval

The percentile bootstrap CI is computed as:

- Lower bound: 2.5th percentile (for 95% CI)
- Upper bound: 97.5th percentile (for 95% CI)

```{r}
#| label: ci
# 95% CI
c(boot_result@ci_lower, boot_result@ci_upper)

# Change confidence level
boot_90 <- bootstrap_mediation(
  med,
  method = "parametric",
  n_boot = 1000,
  ci_level = 0.90
)
c(boot_90@ci_lower, boot_90@ci_upper) # Narrower
```

### Statistical Significance

If the confidence interval excludes zero, the indirect effect is statistically significant:

```{r}
#| label: significance
# Check if CI excludes zero
if (boot_result@ci_lower > 0 || boot_result@ci_upper < 0) {
  print("Indirect effect is statistically significant")
} else {
  print("Indirect effect is not statistically significant")
}
```

## Serial Mediation Bootstrap

Works the same for serial mediation:

```{r}
#| label: serial
# Serial mediation (X -> M1 -> M2 -> Y)
serial_med <- extract_mediation(
  fit_serial,
  treatment = "X",
  mediators = c("M1", "M2"),
  outcome = "Y"
)

# Bootstrap the serial indirect effect (a * d * b)
boot_serial <- bootstrap_mediation(
  serial_med,
  method = "parametric",
  n_boot = 1000,
  ci_level = 0.95
)

# Serial indirect effect
print(boot_serial)

# Point estimate
serial_med@a_path * serial_med@d_path * serial_med@b_path
```

## Reproducibility

Set a seed for reproducible results:

```{r}
#| label: seed
# Same seed = same results
boot1 <- bootstrap_mediation(
  med,
  method = "parametric",
  n_boot = 1000,
  seed = 123
)
boot2 <- bootstrap_mediation(
  med,
  method = "parametric",
  n_boot = 1000,
  seed = 123
)

# Identical results
all.equal(boot1@boot_estimates, boot2@boot_estimates) # TRUE

# Different seed = different results
boot3 <- bootstrap_mediation(
  med,
  method = "parametric",
  n_boot = 1000,
  seed = 456
)
all.equal(boot1@boot_estimates, boot3@boot_estimates) # FALSE
```

## How Many Bootstrap Samples?

General guidelines:

- **Exploratory**: 1000 samples (fast)
- **Publication**: 5000+ samples (stable CI)
- **High stakes**: 10,000+ samples (very stable)

Check stability by varying `n_boot`:

```{r}
#| label: n-boot
# Compare different n_boot
boot_1k <- bootstrap_mediation(
  med,
  method = "parametric",
  n_boot = 1000,
  seed = 123
)
boot_5k <- bootstrap_mediation(
  med,
  method = "parametric",
  n_boot = 5000,
  seed = 123
)
boot_10k <- bootstrap_mediation(
  med,
  method = "parametric",
  n_boot = 10000,
  seed = 123
)

# Compare CIs
rbind(
  "1K" = c(boot_1k@ci_lower, boot_1k@ci_upper),
  "5K" = c(boot_5k@ci_lower, boot_5k@ci_upper),
  "10K" = c(boot_10k@ci_lower, boot_10k@ci_upper)
)
```

If CIs are similar, n_boot is sufficient.

## Alternative CI Methods

medfit currently uses **percentile bootstrap** (simplest, most common).

Future versions may add:

- **BCa (bias-corrected and accelerated)**: Adjusts for bias and skewness
- **Studentized bootstrap**: Better coverage in some cases
- **Bayesian bootstrap**: Posterior intervals

## Comparison with Delta Method (Sobel Test)

The delta method, often called the Sobel test (Sobel, 1982), provides asymptotic standard errors for indirect effects:

### Sobel Standard Error Formula

The first-order delta method approximation yields:

$$SE_{ab} = \sqrt{b^2 \cdot \text{Var}(a) + a^2 \cdot \text{Var}(b)}$$

This is known as the Sobel (1982) formula. Aroian (1947) derived a more accurate formula that includes the product of variances:

$$SE_{ab} = \sqrt{b^2 \cdot \text{Var}(a) + a^2 \cdot \text{Var}(b) + \text{Var}(a) \cdot \text{Var}(b)}$$

```{r}
#| label: delta
# Sobel test SE (first-order approximation)
a <- med@a_path
b <- med@b_path
var_a <- med@vcov["a", "a"]
var_b <- med@vcov["b", "b"]

# Sobel formula
se_sobel <- sqrt(b^2 * var_a + a^2 * var_b)

# Aroian formula (more accurate)
se_aroian <- sqrt(b^2 * var_a + a^2 * var_b + var_a * var_b)

# Sobel 95% CI (assumes normality of ab)
indirect <- a * b
ci_sobel <- indirect + c(-1.96, 1.96) * se_sobel

# Compare with bootstrap
ci_boot <- c(boot_result@ci_lower, boot_result@ci_upper)

rbind(sobel = ci_sobel, bootstrap = ci_boot)
```

### Why Bootstrap is Preferred

MacKinnon et al. (2002) compared the Sobel test to bootstrap methods and found:

1. **Power**: Bootstrap methods have higher statistical power, especially when sample sizes are small or effects are modest
2. **Coverage**: Bootstrap CIs maintain better coverage of the true parameter
3. **Assumptions**: The Sobel test assumes the product $ab$ is normally distributed, which is often violated

The Sobel test tends to be conservative (underpowered) because:

- The product of two normal variables follows a complex distribution, not a normal distribution
- The sampling distribution of $ab$ is typically skewed, especially when $a$ and $b$ are small

Bootstrap methods address these issues by:

- Making no assumptions about the shape of the sampling distribution
- Directly estimating the sampling distribution through resampling
- Providing accurate quantile-based confidence intervals

## Best Practices

1. **Choose method appropriately**:
   - Parametric: Large samples, normal data
   - Nonparametric: Small samples, non-normal data, GLMs
   - Plugin: Quick checks only

2. **Set seed**: Always set seed for reproducibility

3. **Use enough bootstraps**: At least 1000, preferably 5000+

4. **Check convergence**: Ensure models converged on bootstrap samples

5. **Visualize distribution**: Plot histogram and Q-Q plot

6. **Use parallel processing**: For nonparametric with large B

## Next Steps

- See [introduction](introduction.html) for S7 class details
- Learn about [model extraction](extraction.html)
- Check reference documentation for `bootstrap_mediation()`

## Development Status

Phase 5 (Bootstrap Inference) is **complete**:

- ✅ S7 class for BootstrapResult
- ✅ Parametric bootstrap (implemented)
- ✅ Nonparametric bootstrap (implemented)
- ✅ Plugin estimator (implemented)
- ✅ Flexible statistic functions
- ✅ Seed-based reproducibility

See `NEWS.md` for updates.

## References

- Aroian, L. A. (1947). The probability function of the product of two normally distributed variables. *Annals of Mathematical Statistics*, 18(2), 265-271.

- Baron, R. M., & Kenny, D. A. (1986). The moderator-mediator variable distinction in social psychological research: Conceptual, strategic, and statistical considerations. *Journal of Personality and Social Psychology*, 51(6), 1173-1182. https://doi.org/10.1037/0022-3514.51.6.1173

- Craig, C. C. (1936). On the frequency function of xy. *Annals of Mathematical Statistics*, 7(1), 1-15.

- Efron, B. (1979). Bootstrap methods: Another look at the jackknife. *Annals of Statistics*, 7(1), 1-26.

- Efron, B., & Tibshirani, R. J. (1993). *An Introduction to the Bootstrap*. Chapman & Hall/CRC.

- MacKinnon, D. P., Lockwood, C. M., Hoffman, J. M., West, S. G., & Sheets, V. (2002). A comparison of methods to test mediation and other intervening variable effects. *Psychological Methods*, 7(1), 83-104. https://doi.org/10.1037/1082-989X.7.1.83

- Preacher, K. J., & Hayes, A. F. (2008). Asymptotic and resampling strategies for assessing and comparing indirect effects in multiple mediator models. *Behavior Research Methods*, 40(3), 879-891. https://doi.org/10.3758/BRM.40.3.879

- Sobel, M. E. (1982). Asymptotic confidence intervals for indirect effects in structural equation models. *Sociological Methodology*, 13, 290-312. https://doi.org/10.2307/270723
