---
title: "Bootstrap Inference"
format:
  html:
    code-tools: false
    code-copy: false
    highlight-style: none
execute:
  collapse: true
  comment: "#>"
  eval: false
---

## Overview

Bootstrap inference provides confidence intervals for indirect effects without assuming normality. medfit implements three bootstrap methods:

1. **Parametric bootstrap**: Sample from parameter distribution (fast, assumes normality)
2. **Nonparametric bootstrap**: Resample data and refit models (robust, slower)
3. **Plugin estimator**: Point estimate only (fastest, no CI)

All methods return a `BootstrapResult` object with consistent structure.

## Example Data

The examples use the bundled, simulated `mediation_demo` data (see
`?mediation_demo`): a randomized `treatment`, mediators `mediator1` and
`mediator2`, an `outcome`, and two mediator-outcome confounders, `covariate1`
and `covariate2`. Every model below adjusts for both covariates.

```{r}
#| label: setup-data
library(medfit)

data(mediation_demo)
```

## Quick Start: Bootstrap with med()

The easiest way to get bootstrap confidence intervals:

```{r}
#| label: quick-bootstrap
# One-line bootstrap mediation
result <- med(
  data = mediation_demo,
  treatment = "treatment",
  mediator = "mediator1",
  outcome = "outcome",
  covariates = c("covariate1", "covariate2"),
  boot = TRUE,
  n_boot = 1000,
  seed = 123
)

# Instant results with CI
quick(result)
#> NIE = 0.333  [0.217, 0.463] | NDE = 0.206 | PM = 61.8 %

# The BootstrapResult is attached as an attribute (not an S7 property)
attr(result, "bootstrap")
```

For more control over bootstrap methods, use `bootstrap_mediation()` directly.

## Why Bootstrap for Mediation?

The sampling distribution of indirect effects ($a \times b$) is typically:

- **Non-normal**: Even when a and b are normal, their product is not
- **Skewed**: Often right-skewed
- **Complex**: No closed-form distribution for general case

Bootstrap methods:
- Don't assume normality
- Provide accurate coverage (closer to nominal 95%)
- Handle complex indirect effects (serial mediation, moderated mediation)

## Parametric Bootstrap

Samples from the estimated parameter distribution:

```{r}
#| label: parametric
# Fit mediation models
fit_m <- lm(mediator1 ~ treatment + covariate1 + covariate2, data = mediation_demo)
fit_y <- lm(outcome ~ treatment + mediator1 + covariate1 + covariate2,
            data = mediation_demo)

# Extract mediation structure
med_data <- extract_mediation(
  fit_m,
  model_y = fit_y,
  treatment = "treatment",
  mediator = "mediator1"
)

# Define statistic function for indirect effect
# Parameter names are prefixed by model: m_treatment (mediator model),
# y_mediator1 and y_treatment (outcome model)
indirect_effect <- function(theta) {
  theta["m_treatment"] * theta["y_mediator1"]
}

# Parametric bootstrap
boot_result <- bootstrap_mediation(
  statistic_fn = indirect_effect,
  method = "parametric",
  mediation_data = med_data,
  n_boot = 1000,
  ci_level = 0.95,
  seed = 123
)

print(boot_result)
summary(boot_result)
```

### How It Works

1. Extract parameter estimates: $\hat{\theta} = (\hat{a}, \hat{b}, ...)$
2. Extract covariance matrix: $\hat{\Sigma}$
3. For each bootstrap iteration $i = 1, ..., B$:
   - Sample $\theta^*_i \sim N(\hat{\theta}, \hat{\Sigma})$
   - Compute indirect effect: $IE^*_i = a^*_i \times b^*_i$
4. Compute percentile CI from bootstrap distribution

### When to Use

- **Fast**: No model refitting required
- **Appropriate when**: Parameters are approximately normal (large n)
- **Inappropriate when**: Small samples, non-normal parameters

### Advantages

- Very fast (no refitting)
- Reproducible with seed
- Works with extracted models (no need for original data)

### Disadvantages

- Assumes multivariate normality of parameters
- May underestimate uncertainty in small samples

## Nonparametric Bootstrap

Resamples data and refits models:

```{r}
#| label: nonparametric
# Statistic function that refits models on resampled data
statistic_fn_refit <- function(boot_data) {
  # Fit models on bootstrap sample
  fit_m <- lm(mediator1 ~ treatment + covariate1 + covariate2,
              data = boot_data)
  fit_y <- lm(outcome ~ treatment + mediator1 + covariate1 + covariate2,
              data = boot_data)

  # Extract and compute indirect effect
  med_boot <- extract_mediation(fit_m, model_y = fit_y,
                                treatment = "treatment", mediator = "mediator1")
  med_boot@a_path * med_boot@b_path
}

# Nonparametric bootstrap (requires original data)
boot_np <- bootstrap_mediation(
  statistic_fn = statistic_fn_refit,
  method = "nonparametric",
  data = mediation_demo,
  n_boot = 1000,
  ci_level = 0.95,
  seed = 123,
  parallel = TRUE,
  ncores = 4
)

print(boot_np)
```

### How It Works

1. For each bootstrap iteration $i = 1, ..., B$:
   - Resample data with replacement: $D^*_i$
   - Refit mediator model on $D^*_i$
   - Refit outcome model on $D^*_i$
   - Extract paths: $a^*_i$, $b^*_i$
   - Compute indirect effect: $IE^*_i = a^*_i \times b^*_i$
2. Compute percentile CI from bootstrap distribution

### When to Use

- **Robust**: Makes no parametric assumptions
- **Appropriate when**: Small samples, non-normal data, GLMs
- **Gold standard**: Most widely accepted method

### Advantages

- No distributional assumptions
- Robust to outliers
- Captures full sampling variability

### Disadvantages

- Computationally intensive (refits models B times)
- Requires original data
- Can be slow for complex models

### Parallel Processing

Speed up with parallel processing:

```{r}
#| label: parallel
# Detect available cores
ncores <- parallel::detectCores() - 1

# Run in parallel
boot_np_par <- bootstrap_mediation(
  statistic_fn = statistic_fn_refit,
  method = "nonparametric",
  data = mediation_demo,
  n_boot = 5000,
  parallel = TRUE,
  ncores = ncores,
  seed = 123
)
```

## Plugin Estimator

Point estimate only, no confidence interval:

```{r}
#| label: plugin
# Plugin estimator (fastest)
plugin_result <- bootstrap_mediation(
  statistic_fn = indirect_effect,
  method = "plugin",
  mediation_data = med_data
)

print(plugin_result)
```

### When to Use

- Quick checks
- Point estimates for simulation studies
- When CI is not needed

### How It Works

Simply computes $\hat{a} \times \hat{b}$ from the fitted model. No resampling.

## Interpreting Results

### Bootstrap Distribution

The bootstrap distribution shows the sampling variability:

```{r}
#| label: distribution
# Access bootstrap estimates (parametric bootstrap from above)
boot_estimates <- boot_result@boot_estimates

# Histogram
hist(
  boot_estimates,
  breaks = 50,
  main = "Bootstrap Distribution of Indirect Effect",
  xlab = "Indirect Effect (a * b)"
)

# Add percentile CI
abline(v = boot_result@ci_lower, col = "red", lwd = 2)
abline(v = boot_result@ci_upper, col = "red", lwd = 2)
abline(v = boot_result@estimate, col = "blue", lwd = 2)

# Check for normality
qqnorm(boot_estimates)
qqline(boot_estimates)
```

### Confidence Interval

The percentile bootstrap CI is computed as:

- Lower bound: 2.5th percentile (for 95% CI)
- Upper bound: 97.5th percentile (for 95% CI)

```{r}
#| label: ci
# 95% CI
c(boot_result@ci_lower, boot_result@ci_upper)

# Change confidence level
boot_90 <- bootstrap_mediation(
  statistic_fn = indirect_effect,
  method = "parametric",
  mediation_data = med_data,
  n_boot = 1000,
  ci_level = 0.90,
  seed = 123
)
c(boot_90@ci_lower, boot_90@ci_upper) # Narrower
```

### Statistical Significance

If the confidence interval excludes zero, the indirect effect is statistically significant:

```{r}
#| label: significance
# Check if CI excludes zero
if (boot_result@ci_lower > 0 || boot_result@ci_upper < 0) {
  print("Indirect effect is statistically significant")
} else {
  print("Indirect effect is not statistically significant")
}
```

## Serial Mediation Bootstrap

For serial mediation (treatment -> mediator1 -> mediator2 -> outcome), the
indirect effect is the product of the chain's paths, $a \times d \times b$.
The parametric and plugin methods currently accept only a single-mediator
`MediationData` object, so bootstrap a serial chain nonparametrically by
refitting all three models:

```{r}
#| label: serial
# Fit the chain: each model adjusts for the covariates and every upstream
# variable, so the outcome model includes mediator1 as well as mediator2
fit_serial <- function(d) {
  extract_mediation(
    lm(mediator1 ~ treatment + covariate1 + covariate2, data = d),
    model_y = lm(outcome ~ treatment + mediator1 + mediator2 +
                   covariate1 + covariate2, data = d),
    treatment = "treatment",
    mediator = c("mediator1", "mediator2"),
    mediator_models = list(
      lm(mediator2 ~ treatment + mediator1 + covariate1 + covariate2, data = d)
    )
  )
}
serial_med <- fit_serial(mediation_demo)

# Point estimate of the serial indirect effect (a * d * b)
serial_med@a_path * serial_med@d_path * serial_med@b_path

# Bootstrap the serial indirect effect by resampling and refitting
boot_serial <- bootstrap_mediation(
  statistic_fn = function(boot_data) {
    s <- fit_serial(boot_data)
    s@a_path * s@d_path * s@b_path
  },
  method = "nonparametric",
  data = mediation_demo,
  n_boot = 1000,
  ci_level = 0.95,
  seed = 123
)

print(boot_serial)
```

## Reproducibility

Set a seed for reproducible results:

```{r}
#| label: seed
# Same seed = same results
boot1 <- bootstrap_mediation(
  statistic_fn = indirect_effect,
  method = "parametric",
  mediation_data = med_data,
  n_boot = 1000,
  seed = 123
)
boot2 <- bootstrap_mediation(
  statistic_fn = indirect_effect,
  method = "parametric",
  mediation_data = med_data,
  n_boot = 1000,
  seed = 123
)

# Identical results
identical(boot1@boot_estimates, boot2@boot_estimates)
#> [1] TRUE

# Different seed = different results
boot3 <- bootstrap_mediation(
  statistic_fn = indirect_effect,
  method = "parametric",
  mediation_data = med_data,
  n_boot = 1000,
  seed = 456
)
identical(boot1@boot_estimates, boot3@boot_estimates)
#> [1] FALSE
```

## How Many Bootstrap Samples?

General guidelines:

- **Exploratory**: 1000 samples (fast)
- **Publication**: 5000+ samples (stable CI)
- **High stakes**: 10,000+ samples (very stable)

Check stability by varying `n_boot`:

```{r}
#| label: n-boot
# Compare different n_boot
boot_1k <- bootstrap_mediation(
  statistic_fn = indirect_effect,
  method = "parametric",
  mediation_data = med_data,
  n_boot = 1000,
  seed = 123
)
boot_5k <- bootstrap_mediation(
  statistic_fn = indirect_effect,
  method = "parametric",
  mediation_data = med_data,
  n_boot = 5000,
  seed = 123
)
boot_10k <- bootstrap_mediation(
  statistic_fn = indirect_effect,
  method = "parametric",
  mediation_data = med_data,
  n_boot = 10000,
  seed = 123
)

# Compare CIs
rbind(
  "1K" = c(boot_1k@ci_lower, boot_1k@ci_upper),
  "5K" = c(boot_5k@ci_lower, boot_5k@ci_upper),
  "10K" = c(boot_10k@ci_lower, boot_10k@ci_upper)
)
```

If CIs are similar, n_boot is sufficient.

## Alternative CI Methods

medfit currently uses **percentile bootstrap** (simplest, most common).

Future versions may add:

- **BCa (bias-corrected and accelerated)**: Adjusts for bias and skewness
- **Studentized bootstrap**: Better coverage in some cases
- **Bayesian bootstrap**: Posterior intervals

## Comparison with Delta Method

The delta method provides asymptotic SE for indirect effects:

```{r}
#| label: delta
# Delta method SE
a <- med_data@a_path
b <- med_data@b_path
var_a <- vcov(fit_m)["treatment", "treatment"]
var_b <- vcov(fit_y)["mediator1", "mediator1"]
cov_ab <- 0 # Assumed independent

se_delta <- sqrt(b^2 * var_a + a^2 * var_b)

# Delta method 95% CI (assumes normality)
ci_delta <- a * b + c(-1.96, 1.96) * se_delta

# Compare with bootstrap
ci_boot <- c(boot_result@ci_lower, boot_result@ci_upper)

rbind(delta = ci_delta, bootstrap = ci_boot)
```

Bootstrap is generally preferred because it:
- Doesn't assume normality of indirect effect
- Handles skewness correctly
- Provides better coverage

## Best Practices

1. **Choose method appropriately**:
   - Parametric: Large samples, normal data
   - Nonparametric: Small samples, non-normal data, GLMs
   - Plugin: Quick checks only

2. **Set seed**: Always set seed for reproducibility

3. **Use enough bootstraps**: At least 1000, preferably 5000+

4. **Check convergence**: Ensure models converged on bootstrap samples

5. **Visualize distribution**: Plot histogram and Q-Q plot

6. **Use parallel processing**: For nonparametric with large B

## Next Steps

- See [introduction](introduction.html) for S7 class details
- Learn about [model extraction](extraction.html)
- Check reference documentation for `bootstrap_mediation()`

## Working with Bootstrap Results

### Tidyverse Integration

```{r}
#| label: tidy-boot
library(generics)

# Convert to tibble
tidy(boot_result)
#> # A tibble: 1 × 5
#>   term     estimate std.error conf.low conf.high
#>   <chr>       <dbl>     <dbl>    <dbl>     <dbl>
#> 1 estimate    0.333    0.0612    0.217     0.463

# One-row summary
glance(boot_result)
#> # A tibble: 1 × 4
#>   estimate ci_level method     n_boot
#>      <dbl>    <dbl> <chr>       <int>
#> 1    0.333     0.95 parametric   1000
```

### Accessing Results Directly

`BootstrapResult` does not define `coef()` or `confint()` methods; read the
estimate and interval from its properties:

```{r}
#| label: base-boot
# Point estimate
boot_result@estimate

# Confidence interval
c(lower = boot_result@ci_lower, upper = boot_result@ci_upper)

# A different level requires re-running the bootstrap (see above)
c(lower = boot_90@ci_lower, upper = boot_90@ci_upper)
```

## Development Status

Bootstrap infrastructure is **complete**:

- ✅ S7 class for BootstrapResult
- ✅ Parametric bootstrap
- ✅ Nonparametric bootstrap
- ✅ Plugin estimator
- ✅ Parallel processing support
- ✅ Seed-based reproducibility
- ✅ Integration with `med()` via `boot = TRUE`
- ✅ Tidyverse methods (tidy, glance)
- ✅ Direct property access (`@estimate`, `@ci_lower`, `@ci_upper`)
- 📋 BCa confidence intervals (future)

See `NEWS.md` for updates.
