---
title: "medfit Cookbook"
format:
  html:
    code-tools: false
    code-copy: false
    highlight-style: none
execute:
  collapse: true
  comment: "#>"
  eval: true
  echo: true
---

Eleven task-sized recipes. Each says when to use it, runs on `mediation_demo`
(except the cluster and latent-mediator recipes, which simulate their data), and links to the article that explains it. For a
one-page lookup see the [Reference Card](refcard.html).

```{r}
#| label: setup
library(medfit)
library(generics)
covs <- c("covariate1", "covariate2")
```

## 1. One-line mediation

*When:* one mediator, one outcome, a data frame, and you want an answer now.

```{r}
#| label: r1-med
res <- med(mediation_demo, treatment = "treatment", mediator = "mediator1",
           outcome = "outcome", covariates = covs)
quick(res)
```

More: [Getting Started](getting-started.html).

## 2. I already have `lm` fits

*When:* you fitted the mediator and outcome models yourself.

```{r}
#| label: r2-lm
fit_m <- lm(mediator1 ~ treatment + covariate1 + covariate2, mediation_demo)
fit_y <- lm(outcome ~ treatment + mediator1 + covariate1 + covariate2,
            mediation_demo)
x <- extract_mediation(fit_m, model_y = fit_y,
                       treatment = "treatment", mediator = "mediator1")
tidy(x, type = "effects", conf.int = TRUE)
```

More: [Model Extraction](extraction.html).

## 3. Binary outcome

*When:* the outcome is 0/1. Fit the outcome with `glm()`; the `b` path and `c'`
are on the logit scale.

```{r}
#| label: r3-binary
d <- mediation_demo
d$outcome_bin <- as.integer(d$outcome > median(d$outcome))
fit_yb <- glm(outcome_bin ~ treatment + mediator1 + covariate1 + covariate2,
              data = d, family = binomial())
xb <- extract_mediation(fit_m, model_y = fit_yb,
                        treatment = "treatment", mediator = "mediator1")
paths(xb)
```

More: [Model Extraction](extraction.html).

## 4. Weights and robust standard errors

*When:* inverse-probability or survey weights. Pair them with
`se_type = "sandwich"`; model-based standard errors are not valid under
inverse-probability weights.

```{r}
#| label: r4-weights
ps <- glm(treatment ~ covariate1 + covariate2, family = binomial(),
          data = mediation_demo)
p_t <- fitted(ps)
p_m <- mean(mediation_demo$treatment)
ipw <- ifelse(mediation_demo$treatment == 1, p_m / p_t, (1 - p_m) / (1 - p_t))
xw <- fit_mediation(
  outcome ~ treatment + mediator1 + covariate1 + covariate2,
  mediator1 ~ treatment + covariate1 + covariate2,
  data = mediation_demo, treatment = "treatment", mediator = "mediator1",
  weights = ipw, se_type = "sandwich"
)
tidy(xw, type = "effects", conf.int = TRUE)
```

More: [Getting Started](getting-started.html).

## 5. Two mediators in a chain

*When:* treatment, then mediator 1, then mediator 2, then outcome. Put every
mediator in the outcome model.

```{r}
#| label: r5-serial
fit_m1 <- lm(mediator1 ~ treatment + covariate1 + covariate2, mediation_demo)
fit_m2 <- lm(mediator2 ~ treatment + mediator1 + covariate1 + covariate2,
             mediation_demo)
fit_ys <- lm(outcome ~ treatment + mediator1 + mediator2 + covariate1 +
               covariate2, mediation_demo)
xs <- extract_mediation(fit_m1, model_y = fit_ys, treatment = "treatment",
                        mediator = c("mediator1", "mediator2"),
                        mediator_models = list(fit_m2))
c(chain = unname(nie(xs)), every_path = unname(nie(xs, type = "total")),
  total = unname(te(xs)))
```

`nie()` is the full chain `a d b`; `type = "total"` adds the paths that skip a
mediator. More: [Model Extraction](extraction.html).

## 6. Two mediators side by side

*When:* both mediators are driven by the treatment, neither by the other.

```{r}
#| label: r6-parallel
fit_m3 <- lm(mediator3 ~ treatment + covariate1 + covariate2, mediation_demo)
fit_yp <- lm(outcome ~ treatment + mediator1 + mediator3 + covariate1 +
               covariate2, mediation_demo)
xp <- extract_mediation(fit_m1, model_y = fit_yp, treatment = "treatment",
                        mediator = c("mediator1", "mediator3"),
                        mediator_models = list(fit_m3), structure = "parallel")
paths(xp)
nie(xp)
```

The effect is the sum of the per-mediator products. More:
[Model Extraction](extraction.html).

## 7. Treatment-by-mediator interaction

*When:* the outcome model has `treatment * mediator`. You get the four-way
decomposition (CDE, INTref, INTmed, PIE) at a reference mediator level
`m_star`.

```{r}
#| label: r7-interaction
fit_yi <- lm(outcome_int ~ treatment * mediator1 + covariate1 + covariate2,
             mediation_demo)
xi <- extract_mediation(fit_m, model_y = fit_yi, treatment = "treatment",
                        mediator = "mediator1", m_star = 0)
decompose(xi)
```

More: [Model Extraction](extraction.html).

## 8. Many mediators with products

*When:* several mediators and a treatment-by-mediator term. The result is the
joint effect over the mediator block; bootstrap it with `joint_effects()`.

```{r}
#| label: r8-joint
fit_yj <- lm(outcome_int ~ treatment * mediator1 + mediator2 + covariate1 +
               covariate2, mediation_demo)
xj <- extract_mediation(fit_m1, model_y = fit_yj, treatment = "treatment",
                        mediator = c("mediator1", "mediator2"),
                        mediator_models = list(fit_m2))
boot_j <- bootstrap_mediation(
  function(theta) joint_effects(xj, theta)[["nie"]],
  method = "parametric", mediation_data = xj, n_boot = 1000, seed = 1
)
c(estimate = unname(nie(xj)), lower = boot_j@ci_lower, upper = boot_j@ci_upper)
```

More: [Methods and Formulas](methods.html).

## 9. Cluster-randomized trial

*When:* the treatment is assigned to whole clusters (schools) and the mediator
and outcome are measured on individuals. Formulas hold fixed effects only; the
engine adds the random cluster intercept.

```{r}
#| label: r9-cluster-data
#| eval: !expr requireNamespace("lme4", quietly = TRUE)
set.seed(2026)
J <- 40
n <- 10
cl <- rep(seq_len(J), each = n)
x_j <- sample(rep(0:1, length.out = J))
v <- rnorm(J, sd = 0.5)
u <- rnorm(J, sd = 0.5)
cdat <- data.frame(school = factor(cl), X = x_j[cl])
cdat$M <- 0.5 * cdat$X + v[cl] + rnorm(J * n)
mbar <- ave(cdat$M, cdat$school)
cdat$Y <- 0.2 * cdat$X + 0.3 * (cdat$M - mbar) + 0.6 * mbar + u[cl] + rnorm(J * n)
```

```{r}
#| label: r9-cluster-fit
#| eval: !expr requireNamespace("lme4", quietly = TRUE)
xc <- fit_mediation(Y ~ X + M, M ~ X, data = cdat, treatment = "X",
                    mediator = "M", engine = "lmer", cluster = "school")
tidy(xc, type = "effects", conf.int = TRUE)
decompose(xc)
```

With few clusters add `se_type = "kr"` (needs `pbkrtest`), and bootstrap whole
clusters for the product effects:

```{r}
#| label: r9-cluster-boot
#| eval: !expr requireNamespace("lme4", quietly = TRUE)
nie_of <- function(dd) {
  unname(nie(fit_mediation(Y ~ X + M, M ~ X, data = dd, treatment = "X",
                           mediator = "M", engine = "lmer", cluster = "school")))
}
boot_c <- bootstrap_mediation(
  nie_of, method = "nonparametric", data = cdat, cluster = "school",
  n_boot = 200, seed = 123
)
boot_c
```

The warning, when it appears, counts the draws dropped as singular or
non-convergent refits; the interval uses the rest. The own and spillover split
is a large-cluster approximation, and the interval
evidence for the cluster bootstrap is for other designs; see
[Methods and Formulas](methods.html) and [Bootstrap Inference](bootstrap.html).

## 10. I use lavaan

*When:* you fitted the model as a structural equation model. The same
accessors apply.

```{r}
#| label: r10-lavaan
#| eval: !expr requireNamespace("lavaan", quietly = TRUE)
syntax <- "
  mediator1 ~ a * treatment + covariate1 + covariate2
  outcome ~ b * mediator1 + c_prime * treatment + covariate1 + covariate2
"
sem_fit <- lavaan::sem(syntax, data = mediation_demo)
xl <- extract_mediation(sem_fit, treatment = "treatment",
                        mediator = "mediator1", outcome = "outcome")
tidy(xl, type = "effects", conf.int = TRUE)
```

More: [Model Extraction](extraction.html).

## 11. A latent mediator, native engine

*When:* the mediator is a construct measured by several indicators, and you want
its indirect effect without first averaging the indicators into a score. The
mediator `eta` is not a column of the data; the model syntax defines it. The
native engine is experimental.

```{r}
#| label: r11-latent-data
set.seed(11)
n <- 400
X <- rnorm(n)
eta <- 0.5 * X + rnorm(n, sd = 0.9)
ldat <- data.frame(
  X = X,
  m1 = eta + rnorm(n, sd = 0.7),
  m2 = 0.8 * eta + rnorm(n, sd = 0.7),
  m3 = 0.7 * eta + rnorm(n, sd = 0.8)
)
ldat$Y <- 0.4 * eta + 0.2 * X + rnorm(n, sd = 0.8)
```

```{r}
#| label: r11-latent-fit
syntax <- "
  eta =~ m1 + m2 + m3
  eta ~ a * X
  Y ~ b * eta + cp * X
"
xs <- fit_mediation(model = syntax, data = ldat, treatment = "X",
                    mediator = "eta", engine = "native")
tidy(xs, type = "effects", conf.int = TRUE)
```

`fit_sem()` fits the same syntax and returns the `SEMFit`. `mbco_sem()` tests the
indirect effect as the minimum of two likelihood-ratio tests, refitting with
`a == 0` and with `b == 0`:

```{r}
#| label: r11-latent-mbco
mbco_sem(fit_sem(syntax, ldat), "a", "b")
```

More: [Methods and Formulas](methods.html).
