---
title: "Model Extraction"
format: html
execute:
  collapse: true
  comment: "#>"
  eval: false
---

## Overview

The `extract_mediation()` function provides a standardized interface for extracting mediation structures from fitted models. It works with:

- **lm/glm** models (base R) - implemented
- **lavaan** SEM models - implemented
- **lmer** mixed models (future)

Note: OpenMx extraction is planned for a future release.

All extraction methods return a `MediationData` or `SerialMediationData` object, ensuring consistency across modeling frameworks.

## Theoretical Background

### The Baron and Kenny Framework

Modern mediation analysis builds on the foundational work of Baron & Kenny (1986), who formalized the statistical approach to testing mediation hypotheses. In their framework, mediation occurs when a treatment variable X influences an outcome Y through an intervening variable M (the mediator).

The basic mediation model involves three regression equations:

1. **Total effect**: $Y = c \cdot X + e_1$
2. **Mediator model**: $M = a \cdot X + e_2$
3. **Outcome model**: $Y = c' \cdot X + b \cdot M + e_3$

Where:
- $a$ = effect of X on M (the "a path")
- $b$ = effect of M on Y, controlling for X (the "b path")
- $c'$ = direct effect of X on Y, controlling for M (the "c prime" path)
- $c$ = total effect of X on Y

### Effect Decomposition

The total effect decomposes into direct and indirect components:

$$c = c' + ab$$

This identity, known as the **Baron-Kenny decomposition**, means:
- The **indirect effect** ($ab$) represents the portion of the total effect that operates through the mediator
- The **direct effect** ($c'$) represents the portion that operates independently of the mediator

### Parameter Extraction

medfit extracts all parameters needed for effect decomposition and inference:

| Parameter | Description | Source |
|-----------|-------------|--------|
| $a$ | X → M effect | Mediator model coefficient |
| $b$ | M → Y effect (controlling for X) | Outcome model coefficient |
| $c'$ | X → Y direct effect | Outcome model coefficient |
| $\sigma^2_M$ | Residual variance of M | Mediator model |
| $\sigma^2_Y$ | Residual variance of Y | Outcome model |
| $\Sigma$ | Parameter covariance matrix | Combined from both models |

## Extraction Pattern

All extraction methods follow this pattern:

1. Validate inputs (check variable names exist)
2. Extract path coefficients (a, b, c' paths)
3. Extract covariance matrix (for inference)
4. Extract residual variances (if Gaussian)
5. Retrieve data (if available)
6. Create MediationData object
7. Return standardized structure

## Extracting from lm/glm Models

### Simple Mediation

For simple mediation (X -> M -> Y), fit two models:

```{r}
#| label: lm-simple
# Generate example data
set.seed(123)
n <- 200
X <- rnorm(n)
M <- 0.5 * X + rnorm(n, sd = 0.8)
Y <- 0.3 * M + 0.2 * X + rnorm(n, sd = 0.8)
data <- data.frame(X = X, M = M, Y = Y)

# Fit mediation models
model_m <- lm(M ~ X, data = data)
model_y <- lm(Y ~ X + M, data = data)

# Extract mediation structure
med <- extract_mediation(
  model_m,           # Mediator model
  model_y,           # Outcome model
  treatment = "X",
  mediator = "M"
)

# View results
print(med)
summary(med)

# Indirect effect
med@a_path * med@b_path
```

### What Gets Extracted?

The extraction captures:

- **Path coefficients**:
  - `a_path`: Coefficient of X in mediator model (X -> M)
  - `b_path`: Coefficient of M in outcome model (M -> Y, controlling for X)
  - `c_prime`: Coefficient of X in outcome model (direct effect)

- **Full parameter vector**: All coefficients from both models

- **Covariance matrix**: For computing standard errors of functions of parameters

- **Residual variances**: `sigma_m^2` and `sigma_y^2` for Gaussian models

- **Data and metadata**: Sample size, variable names, convergence status

### GLM with Non-Normal Outcomes

Works with any GLM family:

```{r}
#| label: glm-binary
# Binary outcome
Y_binary <- rbinom(n, 1, prob = plogis(0.3 * M + 0.2 * X))
data$Y_binary <- Y_binary

model_m <- lm(M ~ X, data = data)
model_y <- glm(Y_binary ~ X + M, data = data, family = binomial())

med <- extract_mediation(
  model_m,
  model_y,
  treatment = "X",
  mediator = "M"
)

# Note: b_path and c_prime are on logit scale
print(med)
```

### Controlling for Covariates

Include covariates in both models:

```{r}
#| label: covariates
# Add covariates
data$Z1 <- rnorm(n)
data$Z2 <- rnorm(n)

model_m <- lm(M ~ X + Z1 + Z2, data = data)
model_y <- lm(Y ~ X + M + Z1 + Z2, data = data)

med <- extract_mediation(
  model_m,
  model_y,
  treatment = "X",
  mediator = "M"
)

# Path coefficients adjust for Z1 and Z2
print(med)
```

## Extracting from lavaan Models

### Simple Mediation in SEM

```{r}
#| label: lavaan-simple
library(lavaan)

# Define SEM model
model_syntax <- "
  # Mediator model
  M ~ a * X

  # Outcome model
  Y ~ b * M + c_prime * X

  # Indirect effect
  indirect := a * b

  # Total effect
  total := c_prime + a * b
"

# Fit model
fit <- sem(model_syntax, data = data)

# Extract mediation structure
med <- extract_mediation(
  fit,
  treatment = "X",
  mediator = "M",
  outcome = "Y"
)

print(med)
```

### Serial Mediation in SEM

For serial mediation (X -> M1 -> M2 -> Y):

```{r}
#| label: lavaan-serial
# Generate serial mediation data
M1 <- 0.4 * X + rnorm(n, sd = 0.8)
M2 <- 0.5 * M1 + 0.2 * X + rnorm(n, sd = 0.8)
Y_serial <- 0.3 * M2 + 0.1 * M1 + 0.2 * X + rnorm(n, sd = 0.8)
data_serial <- data.frame(X = X, M1 = M1, M2 = M2, Y = Y_serial)

# Define serial mediation model
serial_syntax <- "
  # First mediator
  M1 ~ a * X

  # Second mediator
  M2 ~ d * M1 + X

  # Outcome
  Y ~ b * M2 + M1 + c_prime * X

  # Serial indirect effect (product-of-three)
  serial_indirect := a * d * b
"

fit_serial <- sem(serial_syntax, data = data_serial)

# Extract as SerialMediationData
med_serial <- extract_mediation(
  fit_serial,
  treatment = "X",
  mediators = c("M1", "M2"),
  outcome = "Y"
)

print(med_serial)

# Serial indirect effect
med_serial@a_path * med_serial@d_path * med_serial@b_path
```

## Compatibility with RMediation

The extraction design is compatible with RMediation's lavaan extractor:

```{r}
#| label: rmediation-compat
# RMediation extracts parameters for Distribution of Product method
# medfit extracts the same structure but as S7 classes

# Both packages can work with the same fitted models
# medfit provides infrastructure, RMediation adds DOP/MBCO methods
```

## Error Handling

The extraction uses **checkmate** for fail-fast input validation with informative error messages:
```{r}
#| label: error-handling
# Variable not in model
extract_mediation(
  model_m,
  model_y,
  treatment = "NonExistent",
  mediator = "M"
)
# Error: Assertion on 'treatment in mediator model' failed:
#        Must be element of set {'(Intercept)','X'}, but is 'NonExistent'.

# Wrong type for treatment argument
extract_mediation(
  model_m,
  model_y,
  treatment = 123,  # Should be character
  mediator = "M"
)
# Error: Assertion on 'treatment' failed: Must be of type 'string', not 'double'.

# Mediator not in outcome model
model_y_wrong <- lm(Y ~ X, data = data)  # Missing M
extract_mediation(
  model_m,
  model_y_wrong,
  treatment = "X",
  mediator = "M"
)
# Error: Assertion on 'mediator in outcome model' failed:
#        Must be element of set {'(Intercept)','X'}, but is 'M'.
```

This defensive programming approach catches errors early with clear messages, making debugging easier.

## Advanced Topics

### Extracting Full Parameter Vector

The `estimates` property contains all parameters from both models:

```{r}
#| label: full-params
med <- extract_mediation(model_m, model_y, treatment = "X", mediator = "M")

# All parameters (intercepts + coefficients)
med@estimates

# Access by name
med@estimates["a"]  # a_path
med@estimates["b"]  # b_path
```

### Covariance Matrix for Inference

The covariance matrix is essential for computing standard errors and for parametric bootstrap. medfit constructs a joint covariance matrix from both models.

**Construction Details:**

For separate mediator and outcome models, the parameters from each model are assumed independent (they are estimated from the same data but different response variables). The joint covariance matrix is block-diagonal:

$$\Sigma = \begin{pmatrix} \Sigma_M & 0 \\ 0 & \Sigma_Y \end{pmatrix}$$

where $\Sigma_M$ is the covariance of mediator model parameters and $\Sigma_Y$ is the covariance of outcome model parameters.

```{r}
#| label: vcov
# Covariance matrix of all parameters
vcov_mat <- med@vcov

# Block structure: mediator model params, then outcome model params
dimnames(vcov_mat)

# For Sobel test SE of indirect effect (a*b):
# Since a comes from mediator model and b from outcome model,
# Cov(a,b) = 0 in the block-diagonal structure

a <- med@a_path
b <- med@b_path
var_a <- vcov_mat["a", "a"]
var_b <- vcov_mat["b", "b"]
cov_ab <- vcov_mat["a", "b"]  # Will be 0 for separate models

# Sobel formula (when cov_ab = 0)
se_sobel <- sqrt(b^2 * var_a + a^2 * var_b)

# First-order Taylor approximation (general case with covariance term)
var_indirect <- b^2 * var_a + a^2 * var_b + 2 * a * b * cov_ab
se_indirect <- sqrt(var_indirect)

se_indirect
```

**Note:** For SEM models (lavaan), the covariance matrix is estimated jointly, so $\text{Cov}(a, b)$ may be non-zero. The bootstrap vignette discusses why bootstrap methods are generally preferred over delta-method standard errors.

## Design for Extensibility

The extraction system is designed to accommodate future extensions:

- **New model types**: Add methods for lmer, brms, etc.
- **Complex mediation**: Multiple mediators, moderated mediation
- **Multiple treatments**: Comparative mediation analysis
- **Latent variables**: SEM with measurement models

All methods return the same S7 class structure, ensuring consistency.

## Next Steps

- Learn about [bootstrap inference](bootstrap.html) on extracted models
- See the [introduction](introduction.html) for S7 class details
- Check the reference documentation for `extract_mediation()` methods

## Development Status

Phase 3 (Model Extraction) is **complete**:

- ✅ S7 class definitions (MediationData, SerialMediationData)
- ✅ lm/glm extraction with checkmate validation
- ✅ lavaan extraction with checkmate validation
- 📋 lmer extraction (future)

**Note**: OpenMx extraction has been postponed to a future release.

See `NEWS.md` for updates.

## References

- Baron, R. M., & Kenny, D. A. (1986). The moderator-mediator variable distinction in social psychological research: Conceptual, strategic, and statistical considerations. *Journal of Personality and Social Psychology*, 51(6), 1173-1182. https://doi.org/10.1037/0022-3514.51.6.1173

- MacKinnon, D. P. (2008). *Introduction to Statistical Mediation Analysis*. Lawrence Erlbaum Associates.

- VanderWeele, T. J. (2015). *Explanation in Causal Inference: Methods for Mediation and Interaction*. Oxford University Press.

- Imai, K., Keele, L., & Tingley, D. (2010). A general approach to causal mediation analysis. *Psychological Methods*, 15(4), 309-334. https://doi.org/10.1037/a0020761

- Pearl, J. (2014). Interpretation and identification of causal mediation. *Psychological Methods*, 19(4), 459-481. https://doi.org/10.1037/a0036434
