---
title: "Model Extraction"
format:
  html:
    code-tools: false
    code-copy: false
    highlight-style: none
execute:
  collapse: true
  comment: "#>"
  eval: false
---

## Overview

The `extract_mediation()` function provides a standardized interface for extracting mediation structures from fitted models. It works with:
- **lm/glm** models (base R) - implemented
- **lavaan** SEM models - implemented
- **lmer** mixed models (future)

Extraction returns the S7 class that matches the structure: `MediationData`, `InteractionMediationData`, `SerialMediationData`, `ParallelMediationData`, or `JointMediationData`. All share one interface (`nie()`, `confint()`, `tidy()`, `bootstrap_mediation()`, ...) across modeling frameworks.

**Note:** For quick analysis, consider using `med()` instead - it handles fitting and extraction in one step. Use `extract_mediation()` when you need more control or already have fitted models.

## Quick Comparison: med() vs extract_mediation()

```{r}
#| label: comparison
library(medfit)

# mediation_demo: simulated data bundled with medfit (see ?mediation_demo).
# Its covariates confound the mediator-outcome relation, so adjust for them.
covs <- c("covariate1", "covariate2")

# Simple way: med() does everything
result <- med(data = mediation_demo, treatment = "treatment",
              mediator = "mediator1", outcome = "outcome", covariates = covs)
quick(result)
#> NIE = 0.333  | NDE = 0.206 | PM = 61.8 %

# Advanced way: fit models separately, then extract
fit_m <- lm(mediator1 ~ treatment + covariate1 + covariate2, data = mediation_demo)
fit_y <- lm(outcome ~ treatment + mediator1 + covariate1 + covariate2,
            data = mediation_demo)
result <- extract_mediation(fit_m, model_y = fit_y, treatment = "treatment",
                            mediator = "mediator1")
quick(result)
#> NIE = 0.333  | NDE = 0.206 | PM = 61.8 %
```

## Extraction Pattern

All extraction methods follow this pattern:

1. Validate inputs (check variable names exist)
2. Extract path coefficients (a, b, c' paths)
3. Extract covariance matrix (for inference)
4. Extract residual variances (if Gaussian)
5. Retrieve data (if available)
6. Create MediationData object
7. Return standardized structure

## Extracting from lm/glm Models

### Simple Mediation

For simple mediation (X -> M -> Y), fit two models. Both adjust for the
covariates, which confound the mediator-outcome relation in `mediation_demo`:

```{r}
#| label: lm-simple
# Fit mediation models
model_m <- lm(mediator1 ~ treatment + covariate1 + covariate2,
              data = mediation_demo)
model_y <- lm(outcome ~ treatment + mediator1 + covariate1 + covariate2,
              data = mediation_demo)

# Extract mediation structure
med_data <- extract_mediation(
  model_m,           # Mediator model
  model_y,           # Outcome model
  treatment = "treatment",
  mediator = "mediator1"
)

# View results
print(med_data)
summary(med_data)

# Use effect extractors (recommended)
nie(med_data)   # Indirect effect (a * b)
nde(med_data)   # Direct effect (c')
te(med_data)    # Total effect
pm(med_data)    # Proportion mediated
quick(med_data) # One-line summary
```

### What Gets Extracted?

The extraction captures:

- **Path coefficients**:
  - `a_path`: Coefficient of X in mediator model (X -> M)
  - `b_path`: Coefficient of M in outcome model (M -> Y, controlling for X)
  - `c_prime`: Coefficient of X in outcome model (direct effect)

- **Full parameter vector**: All coefficients from both models

- **Covariance matrix**: For computing standard errors of functions of parameters

- **Residual variances**: `sigma_m^2` and `sigma_y^2` for Gaussian models

- **Data and metadata**: Sample size, variable names, convergence status

### GLM with Non-Normal Outcomes

Works with any GLM family:

```{r}
#| label: glm-binary
# Binary outcome: dichotomize the outcome at its median
demo <- mediation_demo
demo$outcome_bin <- as.integer(demo$outcome > median(demo$outcome))

model_y_bin <- glm(outcome_bin ~ treatment + mediator1 + covariate1 + covariate2,
                   data = demo, family = binomial())

med_bin <- extract_mediation(
  model_m,
  model_y_bin,
  treatment = "treatment",
  mediator = "mediator1"
)

# Note: b_path and c_prime are on logit scale
print(med_bin)
```

### Controlling for Covariates

Every model above includes `covariate1` and `covariate2` in **both**
equations. In `mediation_demo` they affect the mediator and the outcome, so
leaving them out of the outcome model confounds the `b` path:

```{r}
#| label: covariates
# The same models without the covariates
model_m_unadj <- lm(mediator1 ~ treatment, data = mediation_demo)
model_y_unadj <- lm(outcome ~ treatment + mediator1, data = mediation_demo)

med_unadj <- extract_mediation(
  model_m_unadj,
  model_y_unadj,
  treatment = "treatment",
  mediator = "mediator1"
)

# Compare the b paths: adjusted vs unadjusted
round(c(adjusted = med_data@b_path, unadjusted = med_unadj@b_path), 3)
#>   adjusted unadjusted
#>      0.571      0.713
```

### Serial Mediation with lm/glm

For a serial chain (X -> M1 -> M2 -> ... -> Mk -> Y) fit as separate
regressions ("sequential regression"), pass an ordered `mediator` **vector**
plus the mediator models 2..k via `mediator_models`. The first mediator model
(`M1 ~ X`) goes in the usual `object` slot; the outcome model in `model_y`.
The result is a `SerialMediationData` object.

```{r}
#| label: lm-serial
# Serial chain: treatment -> mediator1 -> mediator2 -> outcome
# First mediator model goes in the object slot; the rest in mediator_models
fit_m1 <- lm(mediator1 ~ treatment + covariate1 + covariate2,
             data = mediation_demo)
fit_m2 <- lm(mediator2 ~ treatment + mediator1 + covariate1 + covariate2,
             data = mediation_demo)
# Include every mediator in the outcome model, not only the last one
fit_y  <- lm(outcome ~ treatment + mediator1 + mediator2 + covariate1 + covariate2,
             data = mediation_demo)

med_serial_lm <- extract_mediation(
  fit_m1,
  model_y = fit_y,
  treatment = "treatment",
  mediator = c("mediator1", "mediator2"),  # a length-2 vector selects the serial branch
  mediator_models = list(fit_m2)  # the remaining k minus 1 mediator models
)

# Serial indirect effect: a * d1 * b
med_serial_lm@a_path * med_serial_lm@d_path * med_serial_lm@b_path
```

Include the treatment and every earlier mediator in each downstream model, as
the SEM version below does. Only `mediator2`'s coefficient becomes the `b`
path, but `mediator1` also affects `outcome` directly in `mediation_demo`, so an
outcome model without it (`outcome ~ treatment + mediator2 + ...`) confounds
`b`: `mediator1` is a common cause of `mediator2` and `outcome`. The product
`a * d1 * b` is then the effect through the full chain
`treatment -> mediator1 -> mediator2 -> outcome` only; the direct
`mediator1 -> outcome` path is not part of it.

The order of `mediator_models` is cross-checked against the `mediator` vector:
each model's response and predecessor are validated, so a misordered list
fails fast with an informative error rather than silently producing wrong
`d`-paths.

::: {.callout-important}
## Same data, different CI: lm vs lavaan

An lm/glm serial chain estimates each equation **separately**, so the combined
covariance is **block-diagonal** across chain paths: `cov(a, d_i)`,
`cov(d_i, b)`, and `cov(d_i, d_j)` are all zero by construction. (The
within-outcome-equation covariance, e.g. `cov(b, c')`, is still preserved.)

A single lavaan `sem()` fit of the *same data* estimates all equations jointly
and yields the **full** covariance among chain paths. Because the serial
indirect-effect standard error depends on these off-diagonal covariances, the
CI from an lm chain will generally differ from — and is often tighter than —
the lavaan fit. This is correct given the different estimators; just be aware
that the engine choice changes the interval for identical data.
:::

### Parallel Mediation with lm/glm

In **parallel** mediation the mediators are not chained (X -> M_j -> Y): each
`M_j` is regressed on the treatment (and covariates) only, and all mediators
enter a single outcome model. The total indirect effect is the **sum** of the
per-mediator products, $\sum_j a_j b_j$. The API mirrors the serial case — pass
a `mediator` vector and the remaining mediator models via `mediator_models` — but
set `structure = "parallel"` (or rely on `structure = "auto"`, which infers
parallel when no mediator is regressed on another). The result is a
`ParallelMediationData` object.

```{r}
#| label: lm-parallel
# Parallel mediators mediator1 and mediator3, each driven by the treatment
fit_m1p <- lm(mediator1 ~ treatment + covariate1 + covariate2,
              data = mediation_demo)
fit_m3p <- lm(mediator3 ~ treatment + covariate1 + covariate2,
              data = mediation_demo)
fit_yp  <- lm(outcome ~ treatment + mediator1 + mediator3 + covariate1 + covariate2,
              data = mediation_demo)

med_parallel_lm <- extract_mediation(
  fit_m1p,
  model_y = fit_yp,
  treatment = "treatment",
  mediator = c("mediator1", "mediator3"),
  mediator_models = list(fit_m3p),
  structure = "parallel"
)

# Per-mediator paths and the summed indirect effect
med_parallel_lm@a_paths
med_parallel_lm@b_paths
nie(med_parallel_lm)               # sum_j a_j * b_j
confint(med_parallel_lm, parm = "effects")
```

This model leaves out `mediator2`, a descendant of `mediator1`, so
`mediator1`'s `b` path (about 0.55 in the population) includes its effect
through `mediator2`; see `?mediation_demo` for the reduced-form values.

The mediator equations are fit separately, and the stored `@vcov` sets the
covariances between their coefficients to zero. `cov(a_j, b_{j'})` is exactly
zero for least squares, because the outcome equation contains every regressor
of the mediator equations. `cov(a_j, a_{j'})` is not: when the mediators'
errors are correlated it is nonzero, and the block-diagonal `@vcov` omits it,
so the indirect-effect SE can be off. The `b_j` (and `c'`) share the single
outcome equation, so `cov(b_j, b_{j'})` and `cov(b_j, c')` are preserved. The
`confint()` indirect-effect SE uses the delta method over this `@vcov`, not a
naive per-mediator sum. A lavaan fit, or a nonparametric bootstrap, captures
the omitted covariance; see [Methods and Formulas](methods.html).

### Several Mediators with a Treatment-by-Mediator Product

When the outcome model of a serial or parallel fit carries a treatment-by-mediator
term, `extract_mediation()` returns a `JointMediationData` object with the
**joint** natural effects of the mediators (VanderWeele and Vansteelandt 2014).
The mediators are treated as one block: the NIE runs through every mediator and
every path among them, and it has no per-mediator split. `outcome_int` carries a
`treatment` by `mediator1` product, so it fits the serial chain from above:

```{r}
#| label: lm-joint
fit_yj <- lm(outcome_int ~ treatment * mediator1 + mediator2 +
               covariate1 + covariate2, data = mediation_demo)

med_joint <- extract_mediation(
  fit_m1,                  # the serial mediator1 model from above
  model_y = fit_yj,
  treatment = "treatment",
  mediator = c("mediator1", "mediator2"),
  mediator_models = list(fit_m2)   # the serial mediator2 model
)
med_joint
#> <JointMediationData>
#>   treatment -> {mediator1 -> mediator2} -> outcome_int  (serial mediators, joint effects)
#>   Products: treatment x mediator1 (m* = 0)
#>     mediator1 a* = +0.5826   b = +0.4465   t3 = +0.4906
#>     mediator2 a* = +0.4248   b = +0.2590
#>   c' (t1) = +0.1748   CDE = +0.1748   NDE = +0.2197
#>   Joint NIE (all paths through mediator1, mediator2) = +0.6560
#>   Total = +0.8757   |   n = 400

generics::tidy(med_joint, type = "effects", conf.int = TRUE)
#> # A tibble: 4 × 5
#>   term  estimate std.error conf.low conf.high
#>   <chr>    <dbl>     <dbl>    <dbl>     <dbl>
#> 1 cde      0.175     0.113 -0.0468      0.396
#> 2 nde      0.220     0.116 -0.00710     0.447
#> 3 nie      0.656     0.116  0.428       0.884
#> 4 te       0.876     0.135  0.611       1.14
```

`a*` is each mediator's **total** treatment effect: for `mediator2` it includes the
path through `mediator1` (`a2 + d * a1`). The joint NIE is
`(b1 + t3) * a*_1 + b2 * a*_2`; its population value in `mediation_demo` is
0.585. It is not comparable to the serial `a * d * b` (0.075 on the same data
without the product), which counts only the path through the full chain.

Standard errors use the delta method over a stacked covariance of all the
equations, conditional on the observed covariates (the NDE is evaluated at their
means). For a parametric bootstrap, use `joint_effects()` as the statistic; it
recomputes the effects from any parameter draw:

```{r}
#| label: lm-joint-boot
boot_nie <- bootstrap_mediation(
  function(theta) joint_effects(med_joint, theta)[["nie"]],
  method = "parametric", mediation_data = med_joint, n_boot = 2000, seed = 1
)
c(boot_nie@ci_lower, boot_nie@ci_upper)
#> [1] 0.443 0.896
```

The fit must meet a few requirements, and each violation errors with its cause:
every model carries the same covariates (a medfit limitation), all models are
Gaussian identity-link, unweighted, with an intercept, and fit to the same rows,
every mediator appears in the outcome model, and the treatment is coded 0/1.
Only outcome-model products written with `:` or `*` are supported; a product in
a mediator model, between mediators, with a covariate, or inside `I()` errors. A
product **precomputed as a data column** cannot be recognized from the formula
and would be ignored silently, so write it in the formula. The identification
assumptions are stated for the whole mediator vector; see `?JointMediationData`.

## Extracting from lavaan Models

### Simple Mediation in SEM

```{r}
#| label: lavaan-simple
library(lavaan)

# Define SEM model
model_syntax <- "
  # Mediator model
  mediator1 ~ a * treatment + covariate1 + covariate2

  # Outcome model
  outcome ~ b * mediator1 + c_prime * treatment + covariate1 + covariate2

  # Indirect effect
  indirect := a * b

  # Total effect
  total := c_prime + a * b
"

# Fit model
fit <- sem(model_syntax, data = mediation_demo)

# Extract mediation structure
med_sem <- extract_mediation(
  fit,
  treatment = "treatment",
  mediator = "mediator1",
  outcome = "outcome"
)

print(med_sem)
```

### Serial Mediation in SEM

For serial mediation (treatment -> mediator1 -> mediator2 -> outcome):

```{r}
#| label: lavaan-serial
# Define serial mediation model
serial_syntax <- "
  # First mediator
  mediator1 ~ a * treatment + covariate1 + covariate2

  # Second mediator
  mediator2 ~ d * mediator1 + treatment + covariate1 + covariate2

  # Outcome
  outcome ~ b * mediator2 + mediator1 + c_prime * treatment +
    covariate1 + covariate2

  # Serial indirect effect (product-of-three)
  serial_indirect := a * d * b
"

fit_serial <- sem(serial_syntax, data = mediation_demo)

# Extract as SerialMediationData
med_serial <- extract_mediation(
  fit_serial,
  treatment = "treatment",
  mediator = c("mediator1", "mediator2"),
  outcome = "outcome"
)

print(med_serial)

# Serial indirect effect
med_serial@a_path * med_serial@d_path * med_serial@b_path
```

### Parallel Mediation in SEM

A single `sem()` fit handles parallel mediators directly: regress each mediator
on the treatment and let the outcome depend on all of them. Passing a `mediator`
vector returns a `ParallelMediationData` object (auto-detected, since no mediator
is regressed on another).

```{r}
#| label: lavaan-parallel
# Parallel model: mediator1 and mediator3 each on the treatment; outcome on both
parallel_syntax <- "
  mediator1 ~ a1 * treatment + covariate1 + covariate2
  mediator3 ~ a2 * treatment + covariate1 + covariate2
  outcome   ~ b1 * mediator1 + b2 * mediator3 + c_prime * treatment +
    covariate1 + covariate2

  # Total indirect effect = sum of per-mediator products
  indirect := a1 * b1 + a2 * b2
"

fit_parallel <- sem(parallel_syntax, data = mediation_demo)

med_parallel <- extract_mediation(
  fit_parallel,
  treatment = "treatment",
  mediator = c("mediator1", "mediator3"),
  outcome = "outcome"
)

print(med_parallel)
nie(med_parallel)
```

Unlike the lm/glm engine, the SEM estimates the whole system jointly, so the
extracted `vcov` preserves **all** off-diagonals — including `cov(a_j, b_j)` and
`cov(a_j, a_{j'})`. Standard errors for the indirect effect therefore reflect the
full joint covariance, and (as with serial mediation) the lavaan interval will
generally differ from the lm one for identical data.

## Treatment-Mediator Interaction (Four-Way Decomposition)

When treatment and mediator **interact**, the total effect splits into four
pieces (VanderWeele 2014): the controlled direct effect (CDE), reference
interaction (INTref), mediated interaction (INTmed), and pure indirect effect
(PIE), with `Total = CDE + INTref + INTmed + PIE`, `NDE = CDE + INTref`, and
`NIE = INTmed + PIE`. medfit computes the decomposition; the causal
interpretation (which requires VanderWeele's no-unmeasured-confounding
assumptions) is the analyst's responsibility.

`extract_mediation()` returns an `InteractionMediationData` object whenever the
outcome model carries an `X:M` term. Set the reference mediator level with
`m_star` (default 0).

### Interaction with lm/glm

For lm/glm, the `X:M` term is detected automatically. `mediation_demo`'s
`outcome_int` carries a `treatment` by `mediator1` interaction:

```{r}
#| label: lm-interaction
# Outcome model includes the treatment:mediator1 interaction
fit_mi <- lm(mediator1 ~ treatment + covariate1 + covariate2,
             data = mediation_demo)
fit_yi <- lm(outcome_int ~ treatment * mediator1 + covariate1 + covariate2,
             data = mediation_demo)

med_int <- extract_mediation(
  fit_mi,
  model_y = fit_yi,
  treatment = "treatment",
  mediator = "mediator1",
  m_star = 0           # reference mediator level for the decomposition
)

decompose(med_int)                          # CDE, INTref, INTmed, PIE + effects
confint(med_int, parm = "components")       # delta-method CIs for the components
```

The four-way formulas (continuous `Y` and `M`) are
`CDE = θ₁ + θ₃·m*`, `INTref = θ₃·(E[M|X=0] − m*)`, `INTmed = θ₃·β₁`, and
`PIE = θ₂·β₁`, where `θ₃` is the interaction coefficient and `β₁` the `a` path.
Non-Gaussian outcomes are not yet supported and raise an informative error.

### Interaction with lavaan

In lavaan the interaction enters as a **product variable** (a column you create),
and the model must be fit with `meanstructure = TRUE` so the mediator intercept
(needed for INTref) is estimated. Name the product term via `interaction`:

```{r}
#| label: lavaan-interaction
demo <- mediation_demo
demo$XM <- demo$treatment * demo$mediator1     # product term

fit_int <- sem(
  "mediator1 ~ treatment + covariate1 + covariate2
   outcome_int ~ mediator1 + treatment + XM + covariate1 + covariate2",
  data = demo,
  meanstructure = TRUE
)

med_int_sem <- extract_mediation(
  fit_int,
  treatment = "treatment",
  mediator = "mediator1",
  outcome = "outcome_int",
  interaction = "XM"     # name of the product predictor in the outcome model
)

decompose(med_int_sem)
```

When the interaction is absent (or `decomposition = "two_way"`), extraction falls
back to the standard `MediationData` — so existing two-way workflows are
unchanged.

### Interaction via the regmedint engine

The two paths above extract from models **you** fit. `fit_mediation()` can
instead delegate the fitting to the
[regmedint](https://cran.r-project.org/package=regmedint) package, which
implements VanderWeele's regression-based estimators in closed form, and hand
back the same medfit classes:

```{r}
#| label: regmedint-interaction
# regmedint requires a numeric 0/1 treatment, which mediation_demo has
med_rmi <- fit_mediation(
  formula_y = outcome_int ~ treatment * mediator1 + covariate1 + covariate2,
  formula_m = mediator1 ~ treatment + covariate1 + covariate2,
  data = mediation_demo,
  treatment = "treatment",
  mediator = "mediator1",
  engine = "regmedint"
)

decompose(med_rmi)                        # same four-way layout as above
confint(med_rmi, parm = "components")     # regmedint's analytical SEs
```

The return class follows the same rule as `extract_mediation()`: an `X:M` term
in `formula_y` yields an `InteractionMediationData`, its absence a plain
`MediationData`. Components map from regmedint's own output as
`CDE = cde`, `INTref = pnde − cde`, `INTmed = tnie − pnie`, and `PIE = pnie`.

Standard errors come from regmedint's delta method rather than medfit's, so
they are analytical — no bootstrap needed — and reproduce the SEs
`regmedint::summary()` reports for the same model.

#### The reference mediator level

The four-way split is read off at a reference mediator level $m^*$:

$$\text{CDE} = \theta_1 + \theta_3 m^*$$

$$\text{INTref} = \theta_3\,(E[M \mid X = 0] - m^*)$$

Set it with `fit_mediation(m_star = )`, which defaults to `0` on every engine:

```{r}
#| label: m-star-argument
med_rm_m1 <- fit_mediation(
  formula_y = outcome_int ~ treatment * mediator1 + covariate1 + covariate2,
  formula_m = mediator1 ~ treatment + covariate1 + covariate2,
  data = mediation_demo,
  treatment = "treatment",
  mediator = "mediator1",
  engine = "regmedint",
  m_star = 1
)

med_rm_m1@m_star   # the reference level that was used
med_rm_m1@cde      # direct effect, holding the mediator at that level
```

The two terms shift in exactly compensating directions, so `nde()`, `nie()`,
`te()`, and `pm()` are **invariant** to `m_star` — only the CDE/INTref split
moves. Passing `m_star` to a fit with no treatment-by-mediator term is an
error, not a silent no-op.

The engines reach the same answer by different routes. `engine = "glm"` applies
`m_star` at *extraction* time, after the coefficients are fit;
`engine = "regmedint"` hands the same value to `regmedint::regmedint()` as its
`m_cde` argument, where the closed-form estimator consumes it at *fitting*
time. Because the two names denote one quantity, `m_cde` is not accepted in
`engine_args` — use `m_star`.

#### Other engine settings

Everything regmedint-specific that is *not* the reference level goes through
`engine_args`:

```{r}
#| label: regmedint-args
med_rmi2 <- fit_mediation(
  formula_y = outcome_int ~ treatment * mediator1 + covariate1 + covariate2,
  formula_m = mediator1 ~ treatment + covariate1 + covariate2,
  data = mediation_demo,
  treatment = "treatment",
  mediator = "mediator1",
  engine = "regmedint",
  m_star = 1,           # reference mediator level (default 0)
  engine_args = list(
    c_cond = c(0, 0),   # covariate levels for the conditional effects (one per covariate)
    interaction = TRUE  # override the formula-based auto-detection
  )
)
```

Recognized names are `interaction`, `cvar`, `mreg`, `yreg`, `a0`, `a1`, and
`c_cond`; each replaces a value the adapter would otherwise derive from the
formulas, families, and data.

**Scope.** The regmedint engine needs a numeric 0/1 treatment and a linear
(Gaussian) mediator model; the outcome may be Gaussian or binomial. Outside
that range regmedint's closed-form effects stop being the products of
regression coefficients that `MediationData` and `InteractionMediationData` are
defined in terms of, so the adapter raises an explicit error instead of
returning an object whose numbers would disagree with `nie()`. `weights` and
`se_type = "sandwich"` are likewise refused rather than silently ignored.
`regmedint` is a suggested package — install it with
`install.packages("regmedint")`.

## Compatibility with RMediation

The extraction design is compatible with RMediation's lavaan extractor:

```{r}
#| label: rmediation-compat
# RMediation extracts parameters for Distribution of Product method
# medfit extracts the same structure but as S7 classes

# Both packages can work with the same fitted models
# medfit provides infrastructure, RMediation adds DOP/MBCO methods
```

## Error Handling

The extraction uses **checkmate** for fail-fast input validation with informative error messages:
```{r}
#| label: error-handling
# Variable not in model
extract_mediation(
  model_m,
  model_y,
  treatment = "NonExistent",
  mediator = "mediator1"
)
# Error: Assertion on 'treatment in mediator model' failed: Must be element of
#   set {'(Intercept)','treatment','covariate1','covariate2'}, but is 'NonExistent'.

# Wrong type for treatment argument
extract_mediation(
  model_m,
  model_y,
  treatment = 123,  # Should be character
  mediator = "mediator1"
)
# Error: Assertion on 'treatment' failed: Must be of type 'string', not 'double'.

# Mediator not in outcome model
model_y_wrong <- lm(outcome ~ treatment + covariate1 + covariate2,
                    data = mediation_demo)  # Missing mediator1
extract_mediation(
  model_m,
  model_y_wrong,
  treatment = "treatment",
  mediator = "mediator1"
)
# Error: Assertion on 'mediator in outcome model' failed: Must be element of
#   set {'(Intercept)','treatment','covariate1','covariate2'}, but is 'mediator1'.
```

This defensive programming approach catches errors early with clear messages, making debugging easier.

## Advanced Topics

### Extracting Full Parameter Vector

The `estimates` property contains all parameters from both models:

```{r}
#| label: full-params
med_data <- extract_mediation(model_m, model_y, treatment = "treatment",
                              mediator = "mediator1")

# All parameters (intercepts + coefficients)
med_data@estimates

# Access by name
med_data@estimates["a"]  # a_path
med_data@estimates["b"]  # b_path
```

### Covariance Matrix for Delta Method

Use the covariance matrix for computing standard errors:

```{r}
#| label: vcov
# Covariance matrix of all parameters
vcov_mat <- med_data@vcov

# For delta method SE of indirect effect (a*b):
# Var(ab) = b^2 * Var(a) + a^2 * Var(b) + 2ab*Cov(a,b)

a <- med_data@a_path
b <- med_data@b_path
var_a <- vcov_mat["a", "a"]
var_b <- vcov_mat["b", "b"]
cov_ab <- vcov_mat["a", "b"]

var_indirect <- b^2 * var_a + a^2 * var_b + 2 * a * b * cov_ab
se_indirect <- sqrt(var_indirect)

se_indirect
```

## Design for Extensibility

The extraction system is designed to accommodate future extensions:

- **New model types**: Add methods for lmer, brms, etc.
- **Complex mediation**: Moderated mediation
- **Multiple treatments**: Comparative mediation analysis
- **Latent variables**: SEM with measurement models

All classes share the same interface, so new model types plug into the same effects, intervals and bootstrap.

## Next Steps

- Learn about [bootstrap inference](bootstrap.html) on extracted models
- See the [introduction](introduction.html) for S7 class details
- Read the [formulas behind every effect and standard error](methods.html)
- Check the reference documentation for `extract_mediation()` methods

## Working with Extracted Results

Once you have a `MediationData` object, you can use all medfit functions:

### Effect Extractors

```{r}
#| label: extractors
# Individual effects
nie(med_data)    # Natural Indirect Effect (a * b)
nde(med_data)    # Natural Direct Effect (c')
te(med_data)     # Total Effect (nie + nde)
pm(med_data)     # Proportion Mediated

# All path coefficients
paths(med_data)  # Named vector: a, b, c_prime
```

### Tidyverse Integration

```{r}
#| label: tidy
library(generics)

# Convert to tibble
tidy(med_data)
#> # A tibble: 6 × 3
#>   term    estimate std.error
#>   <chr>      <dbl>     <dbl>
#> 1 a          0.583    0.0989
#> ...

# Just path coefficients or effects
tidy(med_data, type = "paths")
tidy(med_data, type = "effects")

# With confidence intervals
tidy(med_data, conf.int = TRUE)

# One-row summary
glance(med_data)
```

### Base R Methods

```{r}
#| label: base-r
# Standard generics
coef(med_data)                # Path coefficients
coef(med_data, "effects")     # NIE, NDE, TE
vcov(med_data)                # Variance-covariance matrix
confint(med_data)             # 95% confidence intervals
confint(med_data, level = 0.90)
nobs(med_data)                # Sample size
```

## Development Status

Model extraction is **complete**:

- ✅ S7 class definitions (MediationData, InteractionMediationData,
  SerialMediationData, ParallelMediationData, JointMediationData)
- ✅ lm/glm extraction with checkmate validation
- ✅ lavaan extraction with checkmate validation
- ✅ Effect extractors (nie, nde, te, pm, paths)
- ✅ Tidyverse methods (tidy, glance)
- ✅ Base R generics (coef, vcov, confint, nobs)
- 📋 lmer extraction (future)

See `NEWS.md` for updates.
