---
title: "Getting Started with medfit"
format:
  html:
    code-tools: false
    code-copy: false
    highlight-style: none
execute:
  collapse: true
  comment: "#>"
  eval: false
  echo: true
---

## What is medfit?

**medfit** provides unified infrastructure for mediation analysis in R. It offers:

- **ADHD-friendly API** with `med()` for quick analysis and `quick()` for instant results
- **Effect extractors** like `nie()`, `nde()`, `te()`, `pm()`, and `paths()`
- **Tidyverse integration** with `tidy()` and `glance()` methods
- **S7-based classes** for standardized mediation data structures
- **Foundation** for the mediation analysis ecosystem (RMediation, mediationverse)

## Quick Start: The Simplest Way

### One Function: `med()`

```{r}
#| label: med-simple
library(medfit)

# Bundled simulated data (see ?mediation_demo)
data(mediation_demo)

# Run mediation analysis, adjusting for both covariates
result <- med(
  data = mediation_demo,
  treatment = "treatment",
  mediator = "mediator1",
  outcome = "outcome",
  covariates = c("covariate1", "covariate2")
)

# View results
print(result)
```

### Instant Results: `quick()`

```{r}
#| label: quick
# One-line summary
quick(result)
#> NIE = 0.333  | NDE = 0.206 | PM = 61.8 %
```

### Why the Covariates Matter

`mediation_demo` is simulated so that `covariate1` and `covariate2` affect both
the mediators and the outcome. Leaving them out of either model would bias the
mediator-outcome path, so every example in this article adjusts for both. The
data are described in `?mediation_demo`.

### With Bootstrap CI

```{r}
#| label: med-bootstrap
# Get bootstrap confidence intervals
result_boot <- med(
  data = mediation_demo,
  treatment = "treatment",
  mediator = "mediator1",
  outcome = "outcome",
  covariates = c("covariate1", "covariate2"),
  boot = TRUE,
  n_boot = 1000,
  seed = 42
)

quick(result_boot)
#> NIE = 0.333  [0.208, 0.458] | NDE = 0.206 | PM = 61.8 %
```

## Extract Individual Effects

Use dedicated extractor functions:

```{r}
#| label: extractors
# Natural Indirect Effect (a * b)
nie(result)

# Natural Direct Effect (c')
nde(result)

# Total Effect (nie + nde)
te(result)

# Proportion Mediated
pm(result)

# All path coefficients
paths(result)
#> Named num [1:3] 0.583 0.571 0.206
#> - attr(*, "names")= chr [1:3] "a" "b" "c_prime"
```

## Tidyverse Integration

Use `tidy()` and `glance()` for tibble-based workflows:

```{r}
#| label: tidy
library(generics)

# Tidy tibble of all estimates
tidy(result)
#> # A tibble: 6 × 3
#>   term    estimate std.error
#>   <chr>      <dbl>     <dbl>
#> 1 a          0.583   0.0989
#> 2 b          0.571   0.0535
#> 3 c_prime    0.206   0.110
#> 4 nie        0.333    0.0645
#> 5 nde        0.206    0.110
#> 6 te         0.539    0.119

# Just paths with CIs
tidy(result, type = "paths", conf.int = TRUE)

# Just effects
tidy(result, type = "effects")
```

```{r}
#| label: glance
# One-row model summary
glance(result)
#> # A tibble: 1 × 6
#>     nie   nde    te    pm  nobs converged
#>   <dbl> <dbl> <dbl> <dbl> <int> <lgl>
#> 1 0.333 0.206 0.539 0.618   400 TRUE
```

## Base R Methods

Standard R generics work on medfit objects:

```{r}
#| label: base-methods
# Coefficients
coef(result)                    # Path coefficients (a, b, c')
coef(result, type = "effects")  # Effects (nie, nde, te)
coef(result, type = "all")      # All parameters

# Variance-covariance matrix
vcov(result)

# Confidence intervals
confint(result)                       # 95% CI for paths
confint(result, level = 0.90)         # 90% CI
confint(result, parm = "effects")     # delta-method CI for effects

# Number of observations
nobs(result)
```

## Serial Mediation

For serial mediation (treatment -> mediator1 -> mediator2 -> outcome), fit one
model per mediator plus the outcome model, then extract:

```{r}
#| label: serial
fit_m1 <- lm(mediator1 ~ treatment + covariate1 + covariate2, data = mediation_demo)
fit_m2 <- lm(mediator2 ~ treatment + mediator1 + covariate1 + covariate2,
             data = mediation_demo)
# Keep mediator1 in the outcome model: it affects both mediator2 and the
# outcome, so leaving it out would bias the mediator2 -> outcome path
fit_y <- lm(outcome ~ treatment + mediator1 + mediator2 + covariate1 + covariate2,
            data = mediation_demo)

serial_data <- extract_mediation(
  fit_m1,
  model_y = fit_y,
  treatment = "treatment",
  mediator = c("mediator1", "mediator2"),
  mediator_models = list(fit_m2)
)

# Same extractors work
nie(serial_data)   # a * d * b: the effect through the chain mediator1 -> mediator2
nde(serial_data)   # c'
quick(serial_data)
#> [2 mediators] NIE =0.0755| NDE =0.171| PM =30.6%
```

The serial indirect effect covers only the path through both mediators in
order; paths such as mediator1 -> outcome directly are not part of it.

## Advanced Usage

### Extract from Fitted Models

If you already have fitted models:

```{r}
#| label: extract-mediation
# Fit models separately
fit_m <- lm(mediator1 ~ treatment + covariate1 + covariate2, data = mediation_demo)
fit_y <- lm(outcome ~ treatment + mediator1 + covariate1 + covariate2,
            data = mediation_demo)

# Extract mediation structure
med_data <- extract_mediation(
  fit_m,
  model_y = fit_y,
  treatment = "treatment",
  mediator = "mediator1"
)

# All the same methods work
nie(med_data)
tidy(med_data)
quick(med_data)
```

### fit_mediation() for Full Control

```{r}
#| label: fit-mediation
med_data <- fit_mediation(
  formula_y = outcome ~ treatment + mediator1 + covariate1 + covariate2,
  formula_m = mediator1 ~ treatment + covariate1 + covariate2,
  data = mediation_demo,
  treatment = "treatment",
  mediator = "mediator1"
)

# Access path coefficients directly
med_data@a_path  # treatment to mediator
med_data@b_path  # mediator to outcome
med_data@c_prime # treatment to outcome, the direct path
```

### Case Weights and Robust Standard Errors

`weights` passes case weights (for example survey weights or
inverse-probability weights) to both model fits. Model-based standard errors
assume the weights are known precision weights, which inverse-probability
weights are not, so pair them with `se_type = "sandwich"` for
heteroskedasticity-consistent (HC3) standard errors; this needs the suggested
sandwich package. The example builds stabilized inverse-probability weights
for the treatment from the covariates. `mediation_demo`'s treatment is
randomized, so the weights stay close to 1 and mainly illustrate the calls:

```{r}
#| label: weights-sandwich
# Stabilized inverse-probability-of-treatment weights
ps_fit <- glm(treatment ~ covariate1 + covariate2, family = binomial(),
              data = mediation_demo)
p_treat <- fitted(ps_fit)
p_marg  <- mean(mediation_demo$treatment)
ipw <- ifelse(mediation_demo$treatment == 1,
              p_marg / p_treat, (1 - p_marg) / (1 - p_treat))
summary(ipw)
#>    Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
#>  0.7044  0.8933  0.9841  0.9997  1.0954  1.5436

med_ipw <- fit_mediation(
  formula_y = outcome ~ treatment + mediator1 + covariate1 + covariate2,
  formula_m = mediator1 ~ treatment + covariate1 + covariate2,
  data = mediation_demo,
  treatment = "treatment",
  mediator = "mediator1",
  weights = ipw,
  se_type = "sandwich"
)

library(generics)
tidy(med_ipw, type = "effects", conf.int = TRUE)
#> # A tibble: 3 × 5
#>   term  estimate std.error conf.low conf.high
#>   <chr>    <dbl>     <dbl>    <dbl>     <dbl>
#> 1 nie      0.335    0.0657   0.206      0.463
#> 2 nde      0.208    0.115   -0.0180     0.433
#> 3 te       0.542    0.119    0.308      0.776
```

With the default `se_type = "model"` the same fit gives NIE SE 0.0640 and
NDE SE 0.109, and `fit_mediation()` prints a once-per-session message that model-based
standard errors are not valid under inverse-probability weighting. The
weights also apply to the covariate means used by the four-way decomposition
when `formula_y` has a treatment-by-mediator term. The regmedint engine
supports neither `weights` nor `se_type = "sandwich"`.

### Choosing an Engine

`fit_mediation()` defaults to `engine = "glm"`, which fits both models with
`stats::glm()`. A second engine delegates to the suggested
[regmedint](https://cran.r-project.org/package=regmedint) package for
closed-form regression-based effects:

```{r}
#| label: engine-regmedint
# regmedint requires a numeric 0/1 treatment (mediation_demo$treatment is one)
# and the same covariates in both formulas
med_rm <- fit_mediation(
  formula_y = outcome ~ treatment + mediator1 + covariate1 + covariate2,
  formula_m = mediator1 ~ treatment + covariate1 + covariate2,
  data = mediation_demo,
  treatment = "treatment",
  mediator = "mediator1",
  engine = "regmedint"
)

nie(med_rm)   # every generic works the same way
```

If `formula_y` carries a treatment-by-mediator term (for example
`outcome_int ~ treatment * mediator1 + covariate1 + covariate2`), either
engine returns an `InteractionMediationData` with the four-way decomposition —
analytical standard errors in the regmedint case. The reference mediator level
for the CDE is set with `m_star` (default `0`, on both engines); other
engine-specific settings go through `engine_args`. See
[Model Extraction](extraction.html) for the full mapping and its scope limits.

### Custom Bootstrap

```{r}
#| label: bootstrap-custom
# Define custom statistic function
indirect_effect <- function(theta) {
  theta["m_treatment"] * theta["y_mediator1"]
}

# Run bootstrap
boot_result <- bootstrap_mediation(
  statistic_fn = indirect_effect,
  method = "parametric",
  mediation_data = med_data,
  n_boot = 1000,
  ci_level = 0.95,
  seed = 123
)

# Bootstrap results also support tidy/glance
tidy(boot_result)
glance(boot_result)
```

## Learn More

- [Introduction to medfit](introduction.html) - Detailed S7 class documentation
- [Model Extraction](extraction.html) - Extract from lm, glm, lavaan models
- [Bootstrap Inference](bootstrap.html) - Parametric and nonparametric bootstrap
- [Methods and Formulas](methods.html) - Estimands, formulas, and standard errors

## What medfit Covers

- Simple mediation (`MediationData`), with a treatment-by-mediator
  interaction and the four-way decomposition (`InteractionMediationData`)
- Serial and parallel mediators (`SerialMediationData`,
  `ParallelMediationData`), and joint natural effects of several mediators
  with treatment-by-mediator products (`JointMediationData`)
- Extraction from lm/glm and lavaan fits; fitting with the `glm` and
  `regmedint` engines, optional case weights and sandwich standard errors
- Delta-method standard errors in `tidy()` and `confint()`, and parametric,
  nonparametric and plugin bootstrap
- The [Methods and Formulas](methods.html) article gives the math behind
  every effect and standard error

See `NEWS.md` for the latest updates.
