---
title: "Introduction to medfit"
format:
  html:
    code-tools: false
    code-copy: false
    highlight-style: none
execute:
  collapse: true
  comment: "#>"
  eval: false
---

## Overview

**medfit** provides unified infrastructure for mediation analysis in R. It offers:

- **ADHD-friendly API**: `med()` for quick analysis, `quick()` for instant results
- **Effect extractors**: `nie()`, `nde()`, `te()`, `pm()`, `paths()` for mediation effects
- **Tidyverse integration**: `tidy()` and `glance()` methods for tibble workflows
- **Base R generics**: `coef()`, `vcov()`, `confint()`, `nobs()` for S7 classes
- **S7-based classes** for standardized mediation data structures
- **Foundation** for the mediation analysis ecosystem (RMediation, mediationverse)

The package eliminates code duplication across mediation packages by providing shared infrastructure while allowing each package to focus on its unique methodological contributions.

## Core S7 Classes

medfit defines three main S7 classes to represent mediation structures:

### MediationData

Stores simple mediation models (X -> M -> Y):

```{r}
#| label: mediation-data
# Example: Create a MediationData object
med_data <- MediationData(
  a_path = 0.5,           # X -> M effect
  b_path = 0.3,           # M -> Y effect (controlling for X)
  c_prime = 0.2,          # X -> Y direct effect
  treatment = "X",
  mediator = "M",
  outcome = "Y",
  estimates = c(a = 0.5, b = 0.3, c_prime = 0.2),
  vcov = diag(3) * 0.01,  # Covariance matrix
  n_obs = 100L,
  converged = TRUE,
  source_package = "medfit"
)

# Print method shows key information
print(med_data)

# Summary method provides details
summary(med_data)
```

The indirect effect is computed as `a * b`. Use the effect extractors:

```{r}
#| label: indirect-effect
# Effect extractors (recommended)
nie(med_data)   # Natural Indirect Effect (a * b)
nde(med_data)   # Natural Direct Effect (c')
te(med_data)    # Total Effect (nie + nde)
pm(med_data)    # Proportion Mediated
paths(med_data) # All path coefficients

# Direct slot access (advanced)
med_data@a_path
med_data@b_path
med_data@a_path * med_data@b_path
```

### Tidyverse Methods

```{r}
#| label: tidy-methods
library(generics)

# Tidy tibble of estimates
tidy(med_data)
tidy(med_data, type = "paths")    # Just a, b, c'
tidy(med_data, type = "effects")  # Just nie, nde, te

# One-row model summary
glance(med_data)
```

### Base R Methods

```{r}
#| label: base-methods
# Standard generics work on medfit objects
coef(med_data)               # Path coefficients
coef(med_data, "effects")    # nie, nde, te
vcov(med_data)               # Variance-covariance matrix
confint(med_data)            # Confidence intervals
nobs(med_data)               # Number of observations
```

### SerialMediationData

Stores serial mediation models (X -> M1 -> M2 -> ... -> Y):

```{r}
#| label: serial-mediation
# Example: Two-mediator serial mediation (X -> M1 -> M2 -> Y)
serial_data <- SerialMediationData(
  a_path = 0.4,           # X -> M1
  d_path = 0.5,           # M1 -> M2 (scalar for 2 mediators)
  b_path = 0.3,           # M2 -> Y
  c_prime = 0.2,          # X -> Y direct effect
  treatment = "X",
  mediators = c("M1", "M2"),
  outcome = "Y",
  estimates = c(a = 0.4, d = 0.5, b = 0.3, c_prime = 0.2),
  vcov = diag(4) * 0.01,
  mediator_predictors = list(M1 = "X", M2 = c("X", "M1")),
  outcome_predictors = c("X", "M1", "M2"),
  n_obs = 100L,
  converged = TRUE,
  source_package = "medfit"
)

print(serial_data)
```

For two mediators, the serial indirect effect is `a * d * b`:

```{r}
#| label: serial-indirect
# Use extractors (recommended)
nie(serial_data)    # a * d * b
nde(serial_data)    # c'
quick(serial_data)  # One-line summary

# Direct computation
serial_data@a_path * serial_data@d_path * serial_data@b_path
```

**Design for extensibility**: For 3+ mediators, `d_path` becomes a vector:
- 3 mediators: `a * d21 * d32 * b` (product-of-four)
- k mediators: product-of-(k+1)

### BootstrapResult

Stores bootstrap inference results:

```{r}
#| label: bootstrap-result
# Example: Bootstrap result for indirect effect
boot_result <- BootstrapResult(
  estimate = 0.15,        # Point estimate (a * b)
  ci_lower = 0.08,        # Lower CI bound
  ci_upper = 0.25,        # Upper CI bound
  ci_level = 0.95,        # Confidence level
  boot_estimates = rnorm(1000, mean = 0.15, sd = 0.04),  # Bootstrap distribution
  n_boot = 1000L,
  method = "parametric",
  converged = TRUE
)

print(boot_result)
summary(boot_result)
```

## Main Functions

### Quick Start Functions

The simplest way to run mediation analysis:

```{r}
#| label: quick-start
library(medfit)

# Simulate data
set.seed(123)
n <- 200
mydata <- data.frame(X = rnorm(n))
mydata$M <- 0.5 * mydata$X + rnorm(n)
mydata$Y <- 0.3 * mydata$X + 0.4 * mydata$M + rnorm(n)

# One function does it all
result <- med(
  data = mydata,
  treatment = "X",
  mediator = "M",
  outcome = "Y"
)

# Instant summary
quick(result)
#> NIE = 0.19 | NDE = 0.16 | PM = 55%

# With bootstrap CI
result_boot <- med(
  data = mydata,
  treatment = "X",
  mediator = "M",
  outcome = "Y",
  boot = TRUE,
  n_boot = 1000
)

quick(result_boot)
#> NIE = 0.19 [0.08, 0.32] | NDE = 0.16 | PM = 55%
```

### Effect Extractors

Dedicated functions for extracting mediation effects:

```{r}
#| label: effect-extractors
nie(result)   # Natural Indirect Effect (a * b)
nde(result)   # Natural Direct Effect (c')
te(result)    # Total Effect (nie + nde)
pm(result)    # Proportion Mediated
paths(result) # All path coefficients (a, b, c')
```

### extract_mediation()

Extract mediation structure from pre-fitted models:

```{r}
#| label: extract-generic
# From lm/glm models
fit_m <- lm(M ~ X, data = mydata)
fit_y <- lm(Y ~ X + M, data = mydata)

med <- extract_mediation(
  fit_m,
  model_y = fit_y,
  treatment = "X",
  mediator = "M"
)

# From lavaan SEM models
med <- extract_mediation(
  lavaan_fit,
  treatment = "X",
  mediator = "M",
  outcome = "Y"
)
```

See [Model Extraction](extraction.html) for details.

### fit_mediation()

Fit mediation models with formula interface:

```{r}
#| label: fit-generic
# Fit using GLM engine
med <- fit_mediation(
  formula_y = Y ~ X + M,
  formula_m = M ~ X,
  data = mydata,
  treatment = "X",
  mediator = "M"
)
```

### bootstrap_mediation()

Perform bootstrap inference on indirect effects:

```{r}
#| label: bootstrap-generic
# Define statistic function
indirect_fn <- function(theta) {
  theta["m_X"] * theta["y_M"]
}

# Parametric bootstrap
boot_result <- bootstrap_mediation(
  statistic_fn = indirect_fn,
  method = "parametric",
  mediation_data = med,
  n_boot = 1000,
  ci_level = 0.95
)

# tidy/glance work on bootstrap results too
tidy(boot_result)
glance(boot_result)
```

See [Bootstrap Inference](bootstrap.html) for details.

## Package Ecosystem

medfit serves as the foundation for specialized mediation packages:

- **RMediation**: Confidence intervals via distribution methods
  - Uses medfit for extraction
  - Adds Distribution of Product (DOP), MBCO tests

- **mediationverse**: Meta-package for the ecosystem
  - Loads medfit and RMediation together
  - Provides unified documentation

<!-- Future packages (in development):
- **probmed**: Probabilistic mediation effect size (P_med)
- **medrobust**: Sensitivity analysis
-->

## Design Principles

1. **Type safety**: S7 classes with validators ensure data integrity
2. **Defensive programming**: checkmate assertions for fail-fast input validation
3. **Consistency**: Standardized interfaces across model types
4. **Extensibility**: Easy to add new model engines and methods
5. **Minimal dependencies**: Core functionality with minimal external dependencies
6. **Infrastructure focus**: Provides tools, not effect sizes

## Next Steps

- Learn about [model extraction](extraction.html) from different sources
- Explore [bootstrap inference methods](bootstrap.html)
- See the reference documentation for detailed API information

## Development Status

medfit is feature complete. Current status:

- ✅ Phase 2: S7 class architecture (MediationData, SerialMediationData, BootstrapResult)
- ✅ Phase 3: Model extraction (lm/glm, lavaan)
- ✅ Phase 4: Model fitting (`fit_mediation()` with GLM engine)
- ✅ Phase 5: Bootstrap infrastructure (parametric, nonparametric, plugin)
- ✅ Phase 6: Generic functions (`coef`, `vcov`, `confint`, `nobs`, `nie`, `nde`, `te`, `pm`, `paths`, `tidy`, `glance`)
- ✅ Phase 6.5: ADHD-friendly API (`med`, `quick`)
- 🚧 Phase 7: Polish & release

See `NEWS.md` for the latest updates.
