---
title: "Methods and Formulas"
format:
  html:
    code-tools: false
    code-copy: false
    highlight-style: none
execute:
  collapse: true
  comment: "#>"
  eval: true
---

This article collects, in one place, the estimands medfit computes, the
formulas behind every effect, and how each standard error and interval is
obtained. It describes what the code does; the other articles show how to call
it.

medfit computes effects from fitted models. Whether an effect has a causal
interpretation depends on identification assumptions (no unmeasured
confounding of the relevant relations), which medfit does not check. Each
section says which assumptions the estimand needs.

## Notation and conventions

| Symbol | Meaning |
|---|---|
| $X$ | treatment |
| $M$, $M_1, \dots, M_K$ | mediator(s) |
| $Y$ | outcome |
| $C$ | covariates (the same set in every model unless stated) |
| $\bar c$ | sample means of the covariate design columns |
| $\hat\theta$, $\hat\Sigma$ | stacked estimates and their covariance (`@estimates`, `@vcov`) |

Mediator models are $M = \beta_0 + \beta_1 X + \gamma^\top C + e_M$ and the
outcome model is $Y = \theta_0 + \theta_1 X + \theta_2 M + \theta_4^\top C +
e_Y$, with a product term $\theta_3 X M$ where noted. In the path notation
medfit prints, $a = \beta_1$, $b = \theta_2$, and $c' = \theta_1$.

Effects are for a **unit contrast** of the treatment (from $x$ to $x + 1$;
from 0 to 1 when the treatment is binary). The four-way and joint
decompositions use the contrast from 0 to 1 and linear (Gaussian identity-link)
models. The simple, serial and parallel products apply to any continuous or
binary treatment with linear models. A non-identity link (for example a
logistic outcome) puts `b` and `c'` on the link scale, where the product $a b$
is no longer a natural indirect effect on the outcome scale.

## Which object you get

| Fit | Class | Indirect effect |
|---|---|---|
| one mediator, no product | `MediationData` | $a b$ |
| one mediator, $X \times M$ in the outcome | `InteractionMediationData` | $\text{INTmed} + \text{PIE}$ |
| several mediators in a chain, no product | `SerialMediationData` | $a \, d_1 \cdots d_{K-1} \, b$ (chain; `type = "total"` for all paths) |
| several mediators, not chained, no product | `ParallelMediationData` | $\sum_j a_j b_j$ |
| several mediators, $X \times M_i$ in the outcome | `JointMediationData` | joint NIE through all mediators |
| treatment assigned to whole clusters, `lmer` fits | `ClusterMediationData` | $a \, b_B$, split into own $a \, b_W$ and spillover $a (b_B - b_W)$ |

`extract_mediation()` chooses the class from the models; `fit_mediation()`
fits single-mediator models and returns the first two, or, with
`engine = "lmer"` and a `cluster`, the cluster class.

## Simple mediation

With one mediator and no product term,

$$
\text{NIE} = a b, \qquad \text{NDE} = c', \qquad \text{TE} = a b + c',
\qquad \text{PM} = \frac{\text{NIE}}{\text{TE}} .
$$

`pm()` returns `NA` with a warning when $|\text{TE}|$ is numerically zero,
because the ratio is undefined; it has no standard error for the same reason
(bootstrap it instead).

Read causally, NIE and NDE are the natural indirect and direct effects under
no unmeasured treatment–outcome, mediator–outcome and treatment–mediator
confounding given $C$, and no mediator–outcome confounder affected by the
treatment.

## Serial mediation (chain)

For a chain $X \to M_1 \to \cdots \to M_K \to Y$ fit as separate regressions
(each $M_j$ on $X$, the earlier mediators and $C$; the outcome on $X$, all
mediators and $C$), `nie()` reports by default the effect through the
**full chain only**:

$$
\text{NIE}_{\text{chain}} = a \, d_1 \cdots d_{K-1} \, b ,
$$

where $a$ is the $X$ coefficient of $M_1$, $d_j$ the coefficient of $M_j$ in
the model for $M_{j+1}$, and $b$ the coefficient of $M_K$ in the outcome model.
Paths that skip a mediator (for example $X \to M_2 \to Y$, or $M_1 \to Y$
directly) are not part of it. `nie(x, type = "total")` returns the total
indirect effect, the sum over every path from $X$ to $Y$ through at least one
mediator; with two mediators it is $a_1 b_1 + a_2 b_2 + a_1 d \, b_2$, the same
as the joint NIE below when there is no product term.

`te()` is the full total effect, the sum over every directed path from $X$ to
$Y$, including $c'$:

$$
\text{TE} = \big[(I - B)^{-1}\big]_{Y,X}, \qquad
\text{TE} = c' + a_1 b_1 + a_2 b_2 + a_1 d \, b_2 \quad (K = 2),
$$

where $B$ holds the path coefficients among $(X, M_1, \dots, M_K, Y)$. With
linear models and the same covariates in every equation it equals the
treatment coefficient of the outcome regressed on the treatment and covariates
alone. `pm()` is the total indirect effect divided by `te()`. The skip-path
coefficients are stored as `a2`, ..., `b1`, ..., and `d1_3`-style aliases in
`@estimates`, and the delta-method standard error of `te()` differentiates the
whole sum. A path that is not in its model counts as zero; a hand-built object
that lists a path without its coefficient gets `NA` and a warning.

A causal reading of the chain effect needs the simple-mediation assumptions
for every mediator, plus no unmeasured confounding of each mediator–mediator
relation.

## Parallel mediation

With mediators that do not affect each other, each $M_j$ regressed on $X$ and
$C$ and all of them in one outcome model,

$$
\text{NIE} = \sum_{j=1}^{K} a_j b_j , \qquad \text{NDE} = c' .
$$

The sum is the effect through the mediators together. Reading each $a_j b_j$
as the effect through $M_j$ alone additionally requires that the mediators do
not affect one another.

## Treatment–mediator interaction: the four-way decomposition

When the outcome model carries $\theta_3 X M$, `InteractionMediationData`
splits the total effect of a 0/1 treatment into four components (VanderWeele
2014), with reference mediator level $m^*$ (`m_star`, default 0):

$$
\begin{aligned}
\text{CDE} &= \theta_1 + \theta_3 m^* \\
\text{INT}_{\text{ref}} &= \theta_3 \left( E[M \mid X = 0, \bar c] - m^* \right) \\
\text{INT}_{\text{med}} &= \theta_3 \beta_1 \\
\text{PIE} &= \theta_2 \beta_1
\end{aligned}
$$

with $E[M \mid X = 0, \bar c] = \beta_0 + \gamma^\top \bar c$,
$\text{NDE} = \text{CDE} + \text{INT}_{\text{ref}}$,
$\text{NIE} = \text{INT}_{\text{med}} + \text{PIE}$, and
$\text{TE} = \text{NDE} + \text{NIE}$. NDE, NIE and TE do not depend on $m^*$;
only the split of the NDE between CDE and INT<sub>ref</sub> does. The
assumptions are those of simple mediation; the CDE alone needs only no
unmeasured treatment–outcome and mediator–outcome confounding. With
$\theta_3 = 0$ the decomposition reduces to simple mediation. The
`InteractionMediationData` validator checks each identity when the object is
built.

## Several mediators with products: joint natural effects

When a serial or parallel fit has treatment-by-mediator products in the outcome
model,

$$
Y = \theta_0 + \theta_1 X + \sum_i \theta_{2i} M_i
  + \sum_{i \in I} \theta_{3i} X M_i + \theta_4^\top C + e_Y ,
$$

`JointMediationData` reports the **joint** natural effects of the mediators as a
block (VanderWeele and Vansteelandt 2014), where $I$ is the set of mediators
with a product ($\theta_{3i} = 0$ otherwise). The formulas need each mediator's
mean given the treatment and covariates only. For a serial chain, whose models
condition on earlier mediators, medfit obtains these by substitution down the
chain:

$$
\beta^*_{1i} = \beta_{1i} + \sum_{j<i} d_{ij}\, \beta^*_{1j}, \qquad
\mu^*_{0i} = \beta_{0i} + \gamma_i^\top \bar c + \sum_{j<i} d_{ij}\, \mu^*_{0j},
$$

where $d_{ij}$ is the coefficient of $M_j$ in the model for $M_i$
($d_{ij} = 0$ for parallel mediators, so the raw and propagated coefficients
coincide). $\beta^*_{1i}$ is the total effect of the treatment on $M_i$, printed
as `a*`, and $\mu^*_{0i}$ is $E[M_i \mid X = 0, \bar c]$. Then

$$
\begin{aligned}
\text{NIE} &= \sum_i (\theta_{2i} + \theta_{3i})\, \beta^*_{1i} \\
\text{NDE} &= \theta_1 + \sum_{i \in I} \theta_{3i}\, \mu^*_{0i} \\
\text{CDE} &= \theta_1 + \sum_{i \in I} \theta_{3i}\, m^*_i \\
\text{TE} &= \text{NDE} + \text{NIE}
\end{aligned}
$$

`m_star` is a scalar or a vector named by the interacting mediators. With one
mediator these are the four-way NDE, NIE and CDE above.

The NIE is the effect through the mediators together; there is no per-mediator
split, and no part of it is "the effect through $M_1$". The identifying
assumptions are stated for the whole mediator vector: no unmeasured
treatment–outcome, mediators–outcome or treatment–mediators confounding, and no
mediator–outcome confounder affected by the treatment **outside the mediator
vector**. Earlier mediators in a chain are affected by the treatment and
confound later ones; that is allowed because they belong to the vector.

medfit requires every model to carry the same covariates, Gaussian
identity-link unweighted fits with intercepts on the same rows, every mediator
in the outcome model, and mediators listed in causal order; each violation
errors. A product written in the formula (`X * M1`, `X:M1`) is recognized; a
product precomputed as a data column is not, and would be ignored silently.

## Cluster-level treatment: the 2-1-1 design

When the treatment $X_j$ is assigned to whole clusters $j = 1, \dots, J$ (for
example schools) and the mediator $M_{ij}$ and outcome $Y_{ij}$ are measured on
the $n_j$ individuals in each, medfit uses the observed cluster mean
$\bar M_j$ of the mediator over the analysis rows. With cluster-level
covariates $W_j$ and level-1 covariates $C_{ij}$ (entered centered, with their
cluster means),

$$
\begin{aligned}
M_{ij} &= \beta_0 + a X_j + \beta_2^\top C_{ij} + \beta_3^\top W_j + v_j + r_{ij}, \\
Y_{ij} &= \theta_0 + c' X_j + b_W (M_{ij} - \bar M_j) + b_B \bar M_j
          + \theta_4^\top (C_{ij} - \bar C_j) + \theta_5^\top \bar C_j
          + \theta_6^\top W_j + u_j + e_{ij},
\end{aligned}
$$

with random cluster intercepts $v_j$ and $u_j$. The raw parameterization
writes $b_W M_{ij} + \kappa \bar M_j$ instead, so $b_B = b_W + \kappa$ (identity
R1, a medfit-side derivation); medfit finds these terms **by value** (constant
within clusters and an exact affine function of $\bar M_j$, and so on), so
grand-mean centering and `scale()` work, and maps the raw form to the
within form with $\hat\Sigma$ transformed as $J \hat\Sigma J^\top$.

| Effect | Formula | Accessor |
|---|---|---|
| NIE | $a \, b_B$ | `nie()` |
| NDE | $c'$ | `nde()` |
| TE | $a \, b_B + c'$ | `te()` |
| Own-mediator (within) indirect | $a \, b_W$ | `decompose()` |
| Spillover (contextual) indirect | $a (b_B - b_W)$ | `decompose()` |

**The NIE needs no weighting.** Changing $X_j$ shifts every member's mediator by
$a$. The deviations $M_{ij} - \bar M_j$ do not change and $\bar M_j$ moves by
$a$, so $Y$ moves by $a \, b_B$ in every cluster. This holds with a random
slope on the within term, because $X$ is constant within a cluster and
$\bar M_j$ carries no slope (medfit's argument).

**The own/spillover split is an approximation.** Shifting only individual $i$'s
mediator by $a$ moves $\bar M_j$ by $a / n_j$, so with the observed cluster mean
the exact effects in a cluster of $n_j$ members are

$$
\text{own} = a \Big[ b_W + \frac{b_B - b_W}{n_j} \Big], \qquad
\text{spillover} = a (b_B - b_W) \frac{n_j - 1}{n_j},
$$

which add to $a \, b_B$ at every cluster size. Expanding the outcome model in
the members' mediators gives the derivation: member $i$'s own mediator enters
once in the deviation and once in the mean, with coefficient
$b_W + (b_B - b_W) / n_j$, and each of the $n_j - 1$ peers' enters only through
the mean, with coefficient $(b_B - b_W) / n_j$; every mediator moves by $a$.
medfit's tests check the formulas against exact counterfactuals. `decompose()` reports the large-cluster limits
$a \, b_W$ and $a (b_B - b_W)$, labeled the cluster-average, large-cluster
approximation. The own estimate is off by $a (b_B - b_W) / H$ with equal cluster
weights, $H$ the harmonic mean cluster size, and `decompose()` warns when that
gap exceeds half the own effect's standard error; `summary()` always prints it.
Under the peer-mean model of Talloen et al. (2016, p. 364), whose outcome
equation has a coefficient $b_p$ on the member's own mediator and $g$ on the
mean of the other members, a fixed cluster size $N$ gives $b_W = b_p - g / (N - 1)$
and $b_B = b_p + g$. The exact own effect is then $a \, b_p$, Talloen et al.'s
own-mediator effect, while the reported $a \, b_W$ differs from it by
$a g / N$ (not zero).

**Covariance.** Given $(X, M, C, W)$ the outcome errors have mean zero under the
model and $\hat a$ depends only on $(X, M, C, W)$, so $\text{Cov}(\hat a,
\hat b_B)$ and $\text{Cov}(\hat a, \hat c')$ are zero to first order, and
`@vcov` is block-diagonal between the mediator model and the outcome model (a
simulation gate checks the cross-replication correlation of $\hat a$ and
$\hat b_B$). The argument fails under upper-level mediator–outcome
confounding, the same condition that biases the NIE.

**What the estimands need.** In every row the models are linear with no
mediator-by-treatment or mediator-by-covariate products, $X$ is randomized, and
clusters are intact.

| Quantity | Also needs |
|---|---|
| Own $a \, b_W$ | no unmeasured lower-level mediator–outcome confounding; unmeasured upper-level confounders are allowed if additive, **provided level-1 covariates are cluster-mean centered** |
| Spillover, NIE, NDE | also no unmeasured upper-level mediator–outcome confounding; under it the direct and contextual estimators are biased while their sum $c' + a \kappa$ is not |
| TE | nothing beyond randomization |
| The own/spillover split | a cross-world assumption across individuals in a cluster; no treatment-induced mediator–outcome confounding |
| All | interference only through the observed cluster mean, none between clusters; members missing from the analysis rows do not drive their peers' outcomes (the observed cluster mean stands for the realized mean of every member; with a sample of each cluster it is a noisy stand-in and the contextual effect $b_B - b_W$ is biased toward zero, Lüdtke et al. 2008) |

`print()` and `summary()` print this table, with the upper-level claim dropped
when a level-1 covariate has no cluster-mean companion.

**Estimation.** Pass the mediator model and the outcome model, both fitted by
`lme4::lmer()` on the same rows, to `extract_mediation()`, or let
`fit_mediation(engine = "lmer", cluster = )` build them: it drops incomplete
rows once, forms `<M>_cm` and `<M>_cwc` (and the same for each level-1
covariate) on the remaining rows, and fits both models by REML with a random
cluster intercept. Standard errors are the delta-method standard errors of the
section below, from lme4's model-based covariance.

```{r}
#| label: cluster-example
#| eval: !expr requireNamespace("lme4", quietly = TRUE)
library(medfit)
library(generics)
set.seed(2026)
J <- 40
n <- 10
cl <- rep(seq_len(J), each = n)
x_j <- sample(rep(0:1, length.out = J))
v <- rnorm(J, sd = 0.5)
u <- rnorm(J, sd = 0.5)
cdat <- data.frame(school = factor(cl), X = x_j[cl])
cdat$M <- 0.5 * cdat$X + v[cl] + rnorm(J * n)
mbar <- ave(cdat$M, cdat$school)
cdat$Y <- 0.2 * cdat$X + 0.3 * (cdat$M - mbar) + 0.6 * mbar + u[cl] + rnorm(J * n)

fit <- fit_mediation(
  Y ~ X + M, M ~ X, data = cdat, treatment = "X", mediator = "M",
  engine = "lmer", cluster = "school"
)
tidy(fit, type = "effects", conf.int = TRUE)
decompose(fit)
```

With few clusters the model-based standard errors can be too small.
`extract_mediation()` warns when $J < 25$ with model-based standard errors, and
whatever the standard error type when $J < 10$. `se_type = "kr"` replaces each
model's covariance by the Kenward–Roger adjusted one (`pbkrtest::vcovAdj()`,
REML fits only) and stores a Kenward–Roger denominator degrees of freedom for
each path; `confint(parm = "paths")` and `tidy(conf.int = TRUE)` then give $t$
intervals for the paths, while the intervals for the products (NIE, own,
spillover, TE) stay normal. The Kenward–Roger adjustment changes the standard
errors very little (ratios of 1.000 to 1.002 at $J = 15$); its value is the
degrees of freedom. The product intervals are the weak point at small $J$.
`bootstrap_mediation(cluster = )` needs no normal approximation for them; no
few-cluster threshold is established, and the published evidence for resampling
whole clusters is for treatment varying within clusters with 30 or more
clusters (Falk et al. 2024), not for this design. The treatment's effect on the
mediator rests on $J$ cluster-level contrasts however many members each cluster
has, so the number of clusters, not of members, is the sample size for it.

```{r}
#| label: cluster-kr
#| eval: !expr requireNamespace("lme4", quietly = TRUE) && requireNamespace("pbkrtest", quietly = TRUE)
fit_kr <- fit_mediation(
  Y ~ X + M, M ~ X, data = cdat, treatment = "X", mediator = "M",
  engine = "lmer", cluster = "school", se_type = "kr"
)
fit_kr@kr_df
confint(fit_kr)
```

medfit does not support, in this design: treatment varying within clusters
(1-1-1), non-Gaussian outcomes or mediators, treatment-by-mediator or
covariate-by-mediator products, random slopes on anything but the within-cluster
mediator term, or individual-average effects.

## Covariance of the estimates

Every effect's standard error comes from $\hat\Sigma$ (`@vcov`), whose rows are
named by the model coefficients and by path aliases (`a`, `b`, `c_prime`,
`d1`, `a1`, `b2`, `theta3`, …).

**lm/glm engines.** Each equation's block is its model's `vcov()` (or, with
`fit_mediation(se_type = "sandwich")`, the HC3 sandwich estimator
`sandwich::vcovHC(type = "HC3")`, recommended with inverse-probability
`weights`). For ordinary least squares fits of several equations to the same
rows, the covariance between two equations' coefficients is

$$
\operatorname{Cov}(\hat\beta_e, \hat\beta_f) =
\hat\sigma_{ef}\, (X_e^\top X_e)^{-1} X_e^\top X_f\, (X_f^\top X_f)^{-1},
$$

where $\hat\sigma_{ef}$ is the residual cross-product. It is exactly zero when
one equation's residual lies in the other's column space, which holds when the
later equation contains every regressor of the earlier one. So for simple
mediation, serial chains and the four-way model with shared covariates, the
cross-equation blocks are zero and a block-diagonal $\hat\Sigma$ is exact.

- `SerialMediationData`, `ParallelMediationData`, `MediationData` and
  `InteractionMediationData` from lm/glm store a block-diagonal $\hat\Sigma$.
  For parallel mediators whose errors are correlated this omits the
  $\operatorname{Cov}(a_j, a_{j'})$ term, so the NIE standard error can be off.
- `JointMediationData` stores the full stacked covariance above, with
  $\hat\sigma_{ef}$ scaled by $\sqrt{(n - p_e)(n - p_f)}$ so each diagonal
  block equals its model's `vcov()`. Only parallel mediator pairs have non-zero
  cross-blocks. It refuses a non-default `vcov_fun`, because the formula
  assumes OLS.

**lavaan engine.** A single `sem()` fit estimates all equations jointly, so
$\hat\Sigma$ carries every covariance. For the same data its intervals can
differ from the lm/glm ones.

## Delta-method standard errors

For an effect $g(\theta)$, the delta-method variance is
$\nabla g^\top \hat\Sigma \, \nabla g$, evaluated at $\hat\theta$. medfit uses
one gradient builder per class, shared by `confint(parm = "effects")`,
`tidy()` and `summary()`, so they agree exactly. The gradients are analytic:

| Class | Effect | Non-zero partial derivatives |
|---|---|---|
| Simple | NIE $= ab$ | $\partial/\partial a = b$, $\partial/\partial b = a$ |
| Serial | NIE $= a \prod d_j \, b$ | each factor: the product of the others |
| Serial | TE $= [(I - B)^{-1}]_{Y,X}$ | on each path $f \to t$: $[(I - B)^{-1}]_{Y,t} \, [(I - B)^{-1}]_{f,X}$, the sum of the path products through it with it removed; the total NIE drops $c'$ |
| Parallel | NIE $= \sum a_j b_j$ | $\partial/\partial a_j = b_j$, $\partial/\partial b_j = a_j$ |
| all three | NDE $= c'$ | $1$ on $c'$ |
| Simple, parallel | TE | the sum of the NIE and NDE gradients |
| Four-way | CDE, INT<sub>ref</sub>, INT<sub>med</sub>, PIE | in $\theta_1, \theta_2, \theta_3, \beta_0, \beta_1$ and the covariate coefficients (weighted by $\bar c$) |
| Joint | NIE | $\lambda_j$ on $\beta_{1j}$, $\lambda_i \beta^*_{1j}$ on $d_{ij}$, $\beta^*_{1i}$ on $\theta_{2i}$ and $\theta_{3i}$ |
| Joint | NDE | $1$ on $\theta_1$, $\mu^*_{0i}$ on $\theta_{3i}$, $\kappa_j$ on $\beta_{0j}$, $\kappa_j \bar c$ on $\gamma_j$, $\kappa_i \mu^*_{0j}$ on $d_{ij}$ |
| Cluster | NIE $= a b_B$ | $\partial/\partial a = b_B$, $\partial/\partial b_B = a$ |
| Cluster | own $= a b_W$ | $\partial/\partial a = b_W$, $\partial/\partial b_W = a$ |
| Cluster | spillover $= a (b_B - b_W)$ | $b_B - b_W$ on $a$, $a$ on $b_B$, $-a$ on $b_W$ |
| Cluster | NDE, TE | $1$ on $c'$; TE adds the NIE gradient |

For the joint effects the chain rule through the propagated terms is carried
by backward recursions, $\lambda_j = (\theta_{2j} + \theta_{3j}) + \sum_{i>j}
d_{ij} \lambda_i$ and $\kappa_j = \theta_{3j} + \sum_{i>j} d_{ij} \kappa_i$.

The covariate means $\bar c$ are treated as fixed, so standard errors that
involve them (the four-way INT<sub>ref</sub> and the joint NDE) are
conditional on the observed covariates; they slightly understate the
uncertainty of a population-average effect.

**Intervals.** `confint()` and `tidy(conf.int = TRUE)` report normal
intervals, $\hat g \pm z_{1-\alpha/2}\, \widehat{\text{SE}}$. The sampling
distribution of a product of coefficients is skewed, so these can be
inaccurate for indirect effects; `confint()` warns about this for effects,
`tidy()` stays silent. Prefer a bootstrap for inference on an indirect effect.

## Bootstrap

`bootstrap_mediation()` has three methods, each returning a `BootstrapResult`
with the point estimate and a **percentile** interval (the $\alpha/2$ and
$1 - \alpha/2$ quantiles of the bootstrap distribution).

- **Parametric**: draws $\theta^{*(r)} \sim N(\hat\theta, \hat\Sigma)$,
  $r = 1, \dots, R$ (`n_boot`), and applies `statistic_fn` to each named draw.
  Fast; it assumes the estimates are approximately normal and inherits
  $\hat\Sigma$ (including its block structure).
- **Nonparametric**: resamples rows with replacement, refits the models, and
  recomputes the statistic. Slower; it needs no normality assumption and
  captures every covariance, including the covariate means.
- **Plugin**: evaluates the statistic at $\hat\theta$ only, with no interval.
- **Cluster** (`method = "nonparametric"` with `cluster = `): resamples whole
  clusters with replacement and gives each draw a fresh cluster id before the
  refit, so a cluster drawn twice is refit as two clusters, not one larger one.
  A refit that is singular or warns about convergence counts as a failure, is
  excluded, and is counted in the warning.

`statistic_fn` receives the named `@estimates` vector. For the simple, serial
and parallel classes the aliases suffice (for example
`function(theta) theta[["a"]] * theta[["b"]]`). The joint NDE also needs the
intercept and covariate rows and $\bar c$; use `joint_effects()` as the
statistic:

```{r}
#| label: joint-boot-recipe
#| eval: false
bootstrap_mediation(
  function(theta) joint_effects(fit, theta)[["nie"]],
  method = "parametric", mediation_data = fit, n_boot = 2000
)
```

## Fitting engines

`fit_mediation(engine = "glm")` fits the mediator and outcome models with
`glm()` and passes them to `extract_mediation()`, so every formula above
applies. `engine = "regmedint"` delegates to `regmedint::regmedint()`, a
closed-form regression-based implementation, and maps its point estimates and
its own delta-method covariance of the effects onto `MediationData` or
`InteractionMediationData`. The adapter requires a linear mediator model and a
unit treatment contrast; see `?fit_mediation`. For a non-Gaussian outcome
model the components are on that model's link scale, as regmedint reports
them. `engine = "lmer"` fits the two linear mixed models of the cluster design
above (REML, random cluster intercepts, the mediator and level-1 covariates
split into within-cluster deviation and cluster mean) and passes them to
`extract_mediation()`; it returns a `ClusterMediationData`. `engine = "native"`
fits model syntax with the engine described in the next section, then extracts
the mediation structure through the same workers as the lavaan route.

## Native SEM engine

`fit_sem()` and `fit_mediation(engine = "native")` fit a model written in
medfit model syntax (`?"medfit-syntax"`) by maximum likelihood to raw data.
Rows with missing values in the model variables are dropped once (listwise
deletion) and counted.

**Model form.** The model is written in RAM form. With $A$ the matrix of
directed paths, $S_{\text{ram}}$ the matrix of variances and covariances, and
$F$ the selection matrix that picks the observed variables, the implied
covariance of the observed variables is

$$
\Sigma(\theta) = F (I - A)^{-1} S_{\text{ram}} (I - A)^{-\top} F^\top .
$$

**Discrepancy.** With $S$ the sample covariance with divisor $n$ and $p$ the
number of observed variables, the engine minimizes

$$
F_{ML}(\theta) = \log |\Sigma(\theta)| + \operatorname{tr}\left(S \Sigma(\theta)^{-1}\right) - \log |S| - p ,
$$

which is twice lavaan's `fmin`. With the means free,
$-2 \log L = n \left(F_{ML} + \log |S| + p + p \log 2\pi\right)$, and
`logLik()` reports this value. The gradient is analytic. A parameter vector
whose implied covariance is not positive definite gets a large finite value
and a zero gradient, so the optimizer can step back from it.

**Standard errors.** The covariance of $\hat\theta$ is the inverse of the
information for the log likelihood. The default, observed information, is
$(n/2)$ times the central-difference Jacobian of the analytic gradient of
$F_{ML}$ (step $\epsilon^{1/3} \max(|x|, 1)$). Expected information is
$(n/2)\, \operatorname{tr}\left(\Sigma^{-1} \partial_k \Sigma\, \Sigma^{-1} \partial_l \Sigma\right)$.
A singular information matrix gives an `NA` covariance and one warning. Defined
parameters (`:=`) get delta-method standard errors, $\sqrt{g^\top V g}$, with
the gradient $g$ taken by central differences over the labels.

**Constraints.** Equality and inequality constraints that are linear in the
labels become rows $a^\top\theta + b$, normalized to unit length, with constant
Jacobians for the optimizer (SLSQP through nloptr). A nonlinear constraint is
an error. At the solution the active constraints and active bounds define a
null-space basis $Z$ of the feasible directions, and the covariance is
projected,

$$
V = Z \left(Z^\top I Z\right)^{-1} Z^\top ,
$$

so a parameter pinned by a constraint has zero variance and `a == b` returns
the standard error of the model with `a` and `b` collapsed to one label.

**Acceptance.** A solution is accepted only if the optimizer status is a
success code, every constraint holds to $10^{-6}$ in the solver's natural
units, and the Newton decrement $\sqrt{g_r^\top H_r^{-1} g_r}$ on the reduced
gradient and Hessian of the active-constraint null space is at most
$10^{-3}$, with a positive definite reduced Hessian. The decrement is scale
free, so rescaling the data does not change the verdict. The sign of each
active inequality's multiplier is checked and only warns: a wrong sign means
the point may not be a local minimum, and the fit is not rejected for it.

**Starts.** The first start is moment based. If it is not accepted, up to
`n_starts - 1` retries use perturbed starts drawn under a fixed seed derived
from the model, restoring the caller's random number state afterward, so a fit
is reproducible and never changes a simulation around it. When no start is
accepted the error names the best attempt's status and stationarity, and for an
infeasible constraint set, the violated constraint.

```{r}
#| label: native-sem-example
#| echo: true
fit <- fit_sem(
  "mediator1 ~ a*treatment + covariate1
   outcome ~ b*mediator1 + treatment + covariate1
   ab := a*b",
  data = mediation_demo
)
fit@defined
```

## References

Falk, C. F., Vogel, T. A., Hammami, S., & Miočević, M. (2024). Multilevel
mediation analysis in R: A comparison of bootstrap and Bayesian approaches.
*Behavior Research Methods*, 56(2), 750–764.
<https://doi.org/10.3758/s13428-023-02079-4>

Lüdtke, O., Marsh, H. W., Robitzsch, A., Trautwein, U., Asparouhov, T., &
Muthén, B. (2008). The multilevel latent covariate model: A new, more reliable
approach to group-level effects in contextual studies. *Psychological
Methods*, 13(3), 203–229. <https://doi.org/10.1037/a0012869>

Talloen, W., Moerkerke, B., Loeys, T., De Naeghel, J., Van Keer, H., &
Vansteelandt, S. (2016). Estimation of indirect effects in the presence of
unmeasured confounding for the mediator–outcome relationship in a multilevel
2-1-1 mediation model. *Journal of Educational and Behavioral Statistics*,
41(4), 359–391. <https://doi.org/10.3102/1076998616636855>

VanderWeele, T. J. (2010). Direct and indirect effects for neighborhood-based
clustered and longitudinal data. *Sociological Methods & Research*, 38(4),
515–544. <https://doi.org/10.1177/0049124110366236>

VanderWeele, T. J. (2014). A unification of mediation and interaction: A 4-way
decomposition. *Epidemiology*, 25(5), 749–761.

VanderWeele, T. J., & Vansteelandt, S. (2014). Mediation analysis with multiple
mediators. *Epidemiologic Methods*, 2(1), 95–115.
<https://doi.org/10.1515/em-2012-0010>
